---
title: "chaotic_mapping"
format: html
---

```{r}
library(tidyverse)
library(tictoc)
library(metaheuristicOpt)

```



I am going to replace r in the engineWOA with a cubic map ideally. That means for each r_i calculated, i will use f(x_i) where f(x) = map.


```{r}
engineWOA_chaotic <- function(FUN, optimType, maxIter, lowerBound, upperBound, whale) {
  whaleFitness <- calcFitness(FUN, optimType, whale)
  index <- order(whaleFitness)
  whaleFitness <- sort(whaleFitness)
  whale <- whale[index, ]
  bestPos <- whale[1, ]
  FbestPos <- whaleFitness[1]
  curve <- c()
  progressbar <- txtProgressBar(min = 0, max = maxIter, style = 3)

  # chaotic map
  chaotic_map <- function(x) 4 * x^3 - 3 * x
  
  # function to move range from (-1,1) to (0,1)
  to_unit <- function(x) (x + 1) / 2
  
  # starting value for r1, r2 - (-1,1) because that is domain of the map
  r1_val <- runif(1, -1, 1)
  r2_val <- runif(1, -1, 1)
 
  # back to regular function
  for (t in 1:maxIter) {
    a <- 2 - t * ((2) / maxIter)
    a2 <- -1 + t * ((-1) / maxIter)

    for (i in 1:nrow(whale)) {

      # update chaotic values
      r1_val <- chaotic_map(r1_val)
      r2_val <- chaotic_map(r2_val)
      

      # scale to [0,1]
      r1 <- to_unit(r1_val)
      r2 <- to_unit(r2_val)
      

      A <- 2 * a * r1 - a
      C <- 2 * r2
      b <- 1
      p  <- runif(1)
      l  <- (a2 - 1) * runif(1) + 1


      for (j in 1:ncol(whale)) {
        if (p < 0.5) {
          if (abs(A) >= 1) {
            rand.index <- floor(nrow(whale) * to_unit(chaotic_map(r1_val)) + 1)
            x.rand <- whale[rand.index, ]
            D.x.rand <- abs(C * x.rand[j] - whale[i, j])
            whale[i, j] <- x.rand[j] - A * D.x.rand
          } else {
            D.bestPos <- abs(C * bestPos[j] - whale[i, j])
            whale[i, j] <- bestPos[j] - A * D.bestPos
          }
        } else {
          distance <- abs(bestPos[j] - whale[i, j])
          whale[i, j] <- distance * exp(b * l) * cos(l * 2 * pi) + bestPos[j]
        }
      }

      whale[i, ] <- checkBound(whale[i, ], lowerBound, upperBound)
      fitness <- optimType * FUN(whale[i, ])
      if (fitness < FbestPos) {
        FbestPos <- fitness
        bestPos <- whale[i, ]
      }
    }
    curve[t] <- FbestPos
    setTxtProgressBar(progressbar, t)
  }
  close(progressbar)
  curve <- curve * optimType
  return(bestPos)
}

```

Then new WOA function to call that one:

First we need to define this generateRandom(), calcFitness(), checkBound() functions, so I've just copied them from the og package.

```{r}
generateRandom <- function (numPopulation, dimension, lowerBound, upperBound) 
{
    result <- matrix()
    if (length(lowerBound) == 1) {
        result <- matrix(runif(numPopulation * dimension, lowerBound, 
            upperBound), nrow = numPopulation, ncol = dimension)
    }
    else {
        result <- matrix(nrow = numPopulation, ncol = dimension)
        for (i in 1:dimension) {
            result[, i] = runif(numPopulation, lowerBound[i], 
                upperBound[i])
        }
    }
    return(result)
}
```


```{r}
calcFitness <- function (FUN, optimType, popu) 
{
    fitness <- c()
    for (i in 1:nrow(popu)) {
        fitness[i] <- optimType * FUN(popu[i, ])
    }
    return(fitness)
}
```

```{r}
checkBound <- function (position, lowerBound, upperBound) 
{
    check1 <- position > upperBound
    check2 <- position < lowerBound
    result <- (position * (!(check1 + check2))) + upperBound * 
        check1 + lowerBound * check2
    return(result)
}
```



```{r}
WOA_chaotic <- function(FUN, optimType = "MIN", numVar, numPopulation = 40, 
  maxIter = 500, rangeVar) 
{
  dimension <- ncol(rangeVar)
  lowerBound <- rangeVar[1, ]
  upperBound <- rangeVar[2, ]
  if (dimension == 1) {
    dimension <- numVar
  }
  if (optimType == "MAX") 
    optimType <- -1
  else optimType <- 1
  whale <- generateRandom(numPopulation, dimension, lowerBound, 
    upperBound)
  bestPos <- engineWOA_chaotic(FUN, optimType, maxIter, lowerBound, 
    upperBound, whale)
  return(bestPos)
}
```

Now try it on an easy sicegar problem:

First generate data:

```{r}
# generate data
set.seed(47)

# true values
h0 = 30
h1 = 750
t1 = 160
a = 0.05

reps = 5
disp = 250

sig_formula <- function(x, h0, h1, t1, a){
h0 + (h1 - h0) / (1 + exp(-a * (x - t1)))
}

x_vals <- rep(seq(0, 300, by = 20), times = reps)

true_params <- list(h0 = h0, h1 = h1, t1 = t1, a = a)

mean_vals <- rep(sig_formula(x_vals,
                        h0 = true_params$h0,
                        h1 = true_params$h1,
                        t1 = true_params$t1,
                        a = true_params$a))

data <- rnorm(n = length(mean_vals), mean = mean_vals, sd = rep(disp, length(mean_vals)))
sig_data <- data.frame(time = x_vals, intensity = data)

# also want a normalized version
norm_sig_data <- sig_data |>
  mutate(time = ((time - min(time))/(max(time)- min(time))),
         intensity = ((intensity - min(intensity))/(max(intensity)- min(intensity))))

```




```{r}
# objective function is SSE
objective_SSE <- function(x, data) {
  h0 <- x[1]
  h1 <- x[2]
  t1 <- x[3]
  a  <- x[4]
  
  sse <- data |>
    mutate(preds = h0 + (h1 - h0)/(1 + exp(-a*(time - t1))),
           diff = (intensity - preds)^2) |>
    summarize(sse = sum(diff)) |>
    pull()
  
  return(sse)
}

# upper and lower bounds
bounds <- matrix(c(-0.1, 0.3, -0.52, 0.01,
                 0.35, 0.97, 1, 180), nrow = 2, ncol = 4, byrow = TRUE)

# run WOA_chaotic
run_WOA_chaotic <- function(data) {
  # running WOA_chaotic
  sse_result <- WOA_chaotic(
    FUN = function(x) objective_SSE(x, data),
    optimType = "MIN",
    numVar = ncol(bounds),
    numPopulation = 40,
    maxIter = 500,
    rangeVar = bounds
  )
  
  # compute best SSE at that solution
  best_sse <- objective_SSE(sse_result, data)
  
  # return both
  list(
    h0 = sse_result[1],
    h1 = sse_result[2],
    t1 = sse_result[3],
    a  = sse_result[4],
    best_sse = best_sse
  )
}
tic()
run_WOA_chaotic(norm_sig_data)
toc()

```
Some of the time this does as well as the original algorithm. I've loaded in all the functions from WOA, but it still hits errors sometimes, but maybe this is leading to bad estimations (like if some of the helper functions need to be changed. The rest of the time it gives reasonable estimates, but seems to overestimate $a$ most frequently. 

# quasi-random numbers

```{r}
library(randtoolbox)
```

```{r}
engineWOA_quasirandom <- function(FUN, optimType, maxIter, lowerBound, upperBound, whale) {
  whaleFitness <- calcFitness(FUN, optimType, whale)
  index <- order(whaleFitness)
  whaleFitness <- sort(whaleFitness)
  whale <- whale[index, ]
  bestPos <- whale[1, ]
  FbestPos <- whaleFitness[1]
  curve <- c()
  progressbar <- txtProgressBar(min = 0, max = maxIter, style = 3)

  # quasi-random sobol sequence
  dim_needed <- 2   # r1, r2
  sobol_seq <- sobol(n = maxIter, dim = dim_needed, scrambling = 0)

 
  # back to regular function
  for (t in 1:maxIter) {
    a <- 2 - t * ((2) / maxIter)
    a2 <- -1 + t * ((-1) / maxIter)

    for (i in 1:nrow(whale)) {

    r1 <- sobol_seq[t, 1]
    r2 <- sobol_seq[t, 2]
      

      A <- 2 * a * r1 - a
      C <- 2 * r2
      b <- 1
      p  <- runif(1)
      l  <- (a2 - 1) * runif(1) + 1


      for (j in 1:ncol(whale)) {
        if (p < 0.5) {
          if (abs(A) >= 1) {
            rand.index <- floor(nrow(whale) * r1) + 1            
            x.rand <- whale[rand.index, ]
            D.x.rand <- abs(C * x.rand[j] - whale[i, j])
            whale[i, j] <- x.rand[j] - A * D.x.rand
          } else {
            D.bestPos <- abs(C * bestPos[j] - whale[i, j])
            whale[i, j] <- bestPos[j] - A * D.bestPos
          }
        } else {
          distance <- abs(bestPos[j] - whale[i, j])
          whale[i, j] <- distance * exp(b * l) * cos(l * 2 * pi) + bestPos[j]
        }
      }

      whale[i, ] <- checkBound(whale[i, ], lowerBound, upperBound)
      fitness <- optimType * FUN(whale[i, ])
      if (fitness < FbestPos) {
        FbestPos <- fitness
        bestPos <- whale[i, ]
      }
    }
    curve[t] <- FbestPos
    setTxtProgressBar(progressbar, t)
  }
  close(progressbar)
  curve <- curve * optimType
  return(bestPos)
}

```


```{r}
WOA_quasirandom <- function(FUN, optimType = "MIN", numVar, numPopulation = 40, 
  maxIter = 500, rangeVar) 
{
  dimension <- ncol(rangeVar)
  lowerBound <- rangeVar[1, ]
  upperBound <- rangeVar[2, ]
  if (dimension == 1) {
    dimension <- numVar
  }
  if (optimType == "MAX") 
    optimType <- -1
  else optimType <- 1
  whale <- generateRandom(numPopulation, dimension, lowerBound, 
    upperBound)
  bestPos <- engineWOA_quasirandom(FUN, optimType, maxIter, lowerBound, 
    upperBound, whale)
  return(bestPos)
}
```

Test it:

```{r}
bounds <- matrix(c(-0.1, 0.3, -0.52, 0.01,
                 0.35, 0.97, 1, 180), nrow = 2, ncol = 4, byrow = TRUE)

# run WOA_quasirandom
run_WOA_quasirandom <- function(data) {
  # running WOA_quasirandom
  sse_result <- WOA_quasirandom(
    FUN = function(x) objective_SSE(x, data),
    optimType = "MIN",
    numVar = ncol(bounds),
    numPopulation = 40,
    maxIter = 500,
    rangeVar = bounds
  )
  
  # compute best SSE at that solution
  best_sse <- objective_SSE(sse_result, data)
  
  # return both
  list(
    h0 = sse_result[1],
    h1 = sse_result[2],
    t1 = sse_result[3],
    a  = sse_result[4],
    best_sse = best_sse
  )
}
tic()
run_WOA_quasirandom(norm_sig_data)
toc()
```
# testing function

I want to map different datasets accross the three to see how they do. This is using 250 dispersion.

```{r}
data_gen_func <- function(h0 = 30, h1 = 750, t1 = 160, a = 0.05, reps = 5, disp = 250){
  

sig_formula <- function(x, h0, h1, t1, a){
h0 + (h1 - h0) / (1 + exp(-a * (x - t1)))
}

x_vals <- rep(seq(0, 300, by = 20), times = reps)

true_params <- list(h0 = h0, h1 = h1, t1 = t1, a = a)

mean_vals <- rep(sig_formula(x_vals,
                        h0 = true_params$h0,
                        h1 = true_params$h1,
                        t1 = true_params$t1,
                        a = true_params$a))

data <- rnorm(n = length(mean_vals), mean = mean_vals, sd = rep(disp, length(mean_vals)))
sig_data <- data.frame(time = x_vals, intensity = data)

# also want a normalized version
norm_sig_data <- sig_data |>
  mutate(time = ((time - min(time))/(max(time)- min(time))),
         intensity = ((intensity - min(intensity))/(max(intensity)- min(intensity))))
}
```

Quickly define `run_WOA` function.

```{r}
bounds <- matrix(c(-0.1, 0.3, -0.52, 0.01,
                 0.35, 0.97, 1, 180), nrow = 2, ncol = 4, byrow = TRUE)

# run WOA_quasirandom
run_WOA <- function(data) {
  # running WOA
  sse_result <- WOA(
    FUN = function(x) objective_SSE(x, data),
    optimType = "MIN",
    numVar = ncol(bounds),
    numPopulation = 40,
    maxIter = 500,
    rangeVar = bounds
  )
  
  # compute best SSE at that solution
  best_sse <- objective_SSE(sse_result, data)
  
  # return both
  list(
    h0 = sse_result[1],
    h1 = sse_result[2],
    t1 = sse_result[3],
    a  = sse_result[4],
    best_sse = best_sse
  )
}
tic()
run_WOA(norm_sig_data)
toc()
```

In the code below, since future_map() operates in multiple R sessions, we have to be really careful with the random number generation, because they might not be random relative to each other. The warning I kept getting was:

Warning: UNRELIABLE VALUE: Future (NULL) unexpectedly generated random numbers without specifying argument 'seed'. There is a risk that those random numbers are not statistically sound and the overall results might be invalid. To fix this, specify 'seed=TRUE'. This ensures that proper, parallel-safe random numbers are produced. To disable this check, use 'seed=NULL', or set option 'future.rng.onMisuse' to "ignore".

The warning is because the various WOA functions randomly generate numbers internally and the `set.seed()` doesn't flow through the whole process. So the different workers might generate random numbers from the same sequences. I solved this by using the function `withSeed()` from the R.utils package. This sets the seed for both the map and down through the runifs in WOA. I also have the argument `seed = TRUE` in each function which should ensure independence in the random number generation in each function

```{r}
library(furrr)
library(R.utils)

plan(multisession, workers = 5)  # i have 8 total

seeds <- 1:250 

# generate the data with given seeds
generated_data <- map(seeds, function(s) {
  set.seed(s)
  data_gen_func()
})

# define bounds
bounds <- matrix(c(-0.1, 0.3,
                   -0.52, 0.01,
                   0.35, 0.97,
                   1, 180),
                 nrow = 2, ncol = 4, byrow = TRUE)

# helper function to run everything smoothly in parallel
run_woa_parallel <- function(data, i, algo_func, algo_name) {
  out <- withSeed({
    algo_func(data)
  }, seed = seeds[i])  # ensures all random numbers inside algo_func are reproducible

  out$dataset_id <- i
  out$seed <- seeds[i]
  out$algorithm <- algo_name
  out
}

# og WOA
og_df <- future_map2_dfr(
  generated_data, seq_along(generated_data),
  ~ run_woa_parallel(.x, .y, run_WOA, "WOA"),
  seed = TRUE  # ensures parallel-safe independent RNG streams
)

# chaotic WOA
chaotic_df <- future_map2_dfr(
  generated_data, seq_along(generated_data),
  ~ run_woa_parallel(.x, .y, run_WOA_chaotic, "WOA_chaotic"),
  seed = TRUE
)

# quasirandom WOA
quasi_df <- future_map2_dfr(
  generated_data, seq_along(generated_data),
  ~ run_woa_parallel(.x, .y, run_WOA_quasirandom, "WOA_quasirandom"),
  seed = TRUE
)

# combine
all_results <- bind_rows(og_df, chaotic_df, quasi_df)

```

# LMA

I should LMA on the same 250 datasets, then cbind it to all_results.

```{r}
# first define SSE function
sse_func <- function(h0, h1, t1, a, data) {
  sse <- data |>
    mutate(preds = h0 + (h1 - h0) / (1 + exp(-a * (x - t1))),
      diff = (y - preds)^2) |>
    summarize(sse = sum(diff)) |>
    pull()
  return(sse) # i'd rather this be a vector, but this only way i can get it to work down below
}

# and sig_formula
sig_formula <- function(x, h0, h1, t1, a){
h0 + (h1 - h0) / (1 + exp(-a * (x - t1)))
}
```


```{r}
library(minpack.lm)
library(dplyr)
library(purrr)
library(tibble)

# Define the fitting function
run_lma_fit <- function(data, dataset_id, seed) {
  # define bounds & starting values
   startlist <- list(h0 = 0.15, h1 = 0.75, t1 = 0.25, a = 90)
   lowerbounds <- c(h0 = -0.1, h1 = 0.3, t1 = -52, a = 0.01)
   upperbounds <- c(h0 = 0.35,  h1 = 0.97, t1 = 1, a = 180)

  # safely run LMA in case of convergence errors
  safe_nlsLM <- purrr::safely(minpack.lm::nlsLM, otherwise = NULL)
  
  result <- safe_nlsLM(
    intensity ~ sig_formula(time, h0, h1, t1, a),
    data = data,
    start = startlist,
    control = list(maxiter = 1000, factor = 100, ptol = 1e-8, ftol = 1e-8),
    lower = lowerbounds,
    upper = upperbounds,
    trace = FALSE
  )
  
  # If it failed, return NA row
  if (is.null(result$result)) {
    return(tibble(
      dataset_id = dataset_id,
      seed = seed,
      h0 = NA, h1 = NA, t1 = NA, a = NA,
      best_sse = NA,
      algorithm = "LMA"
    ))
  }
  
  # extract estimated parameters
  params <- coef(result$result)
  
  # compute SSE
  sse_val <- sse_func(
    h0 = params["h0"],            
    h1 = params["h1"],
    t1 = params["t1"],
    a  = params["a"],
    data = data |> rename(x = time, y = intensity)
  )
  
  # return as tibble
  tibble(
    h0 = params["h0"],
    h1 = params["h1"],
    t1 = params["t1"],
    a  = params["a"],
    best_sse = sse_val,
    dataset_id = dataset_id,
    seed = seed,
    algorithm = "LMA"
  )
}

tic()
# Run LMA fits on all datasets
lma_results_df <- map2_dfr(generated_data, seeds, ~run_lma_fit(.x, dataset_id = which(seeds == .y), seed = .y))
toc()

# View results
print(lma_results_df)

```

```{r}
# now i want to combine the dataframes with WOA and LMA
all_results <- rbind(all_results, lma_results_df)

```



 true values are h0 = 0.1386869, h1 = 0.8337327, t1 = 0.5333333, a = 15


```{r}
all_results |>
  ggplot(aes(x = h0, y = t1)) +
  geom_point(aes(color = algorithm)) +
  geom_hline(yintercept = 0.53, color = "red") +
  geom_vline(xintercept = 0.15, color = "red") +
  facet_wrap(~algorithm)
```

Almost always overestimates h0 - even looks like it's hitting the upperbound pretty frequently. t1 seems like it is reasonably distributed around the truth. Also, all three algorithms look really similar.

```{r}
all_results |>
  filter(h0 == 0.35) |>
  summarise(h0_hit_upperbound = n())

121/750
```


```{r}
all_results |>
  ggplot(aes(x = a, y = h1)) +
  geom_point(aes(color = algorithm)) +
  geom_hline(yintercept = 0.83, color = "red") +
  geom_vline(xintercept = 15, color = "red") +
  facet_wrap(~algorithm)
```

h1 is consistently underestimated, except for when a is underestimated (interesting). Most of the estimations for a look pretty good.

```{r}
lowest_sse_df <- all_results |>
  group_by(dataset_id) |>
  mutate(lowest_sse = min(best_sse)) |>
  filter(best_sse == lowest_sse) |>
  select(dataset_id, algorithm, lowest_sse)

lowest_sse_df |>
  group_by(algorithm) |>
  summarise(count = n())

```

## dispersion 400

```{r}
library(furrr)
library(R.utils)

seeds <- 1:250 

# generate the data with given seeds
generated_data_400 <- map(seeds, function(s) {
  set.seed(s)
  data_gen_func(disp = 400) # change dispersion here
})

# define bounds
bounds <- matrix(c(-0.1, 0.3,
                   -0.52, 0.01,
                   0.35, 0.97, # this should maybe be 0.3 lower bound for h1
                   1, 180),
                 nrow = 2, ncol = 4, byrow = TRUE)

# helper function to run everything smoothly in parallel
run_woa_parallel <- function(data, i, algo_func, algo_name) {
  out <- withSeed({
    algo_func(data)
  }, seed = seeds[i])  # ensures all random numbers inside algo_func are reproducible

  out$dataset_id <- i
  out$seed <- seeds[i]
  out$algorithm <- algo_name
  out
}

# og WOA
og_df_400 <- future_map2_dfr(
  generated_data_400, seq_along(generated_data_400),
  ~ run_woa_parallel(.x, .y, run_WOA, "WOA"),
  seed = TRUE  # ensures parallel-safe independent RNG streams
)

# chaotic WOA
chaotic_df_400 <- future_map2_dfr(
  generated_data_400, seq_along(generated_data_400),
  ~ run_woa_parallel(.x, .y, run_WOA_chaotic, "WOA_chaotic"),
  seed = TRUE
)

# quasirandom WOA
quasi_df_400 <- future_map2_dfr(
  generated_data_400, seq_along(generated_data_400),
  ~ run_woa_parallel(.x, .y, run_WOA_quasirandom, "WOA_quasirandom"),
  seed = TRUE
)

# combine
all_results_400 <- bind_rows(og_df_400, chaotic_df_400, quasi_df_400)

```

Now with LMA:

```{r}
tic()
# Run LMA fits on all datasets
lma_results_df_400 <- map2_dfr(generated_data_400, seeds, ~run_lma_fit(.x, dataset_id = which(seeds == .y), seed = .y))
toc()

# View results
print(lma_results_df_400)
```

```{r}
# now combine the LMA df with the WOA df
all_results_400 <- rbind(all_results_400, lma_results_df_400)
```




true values are h0 = 0.1386869, h1 = 0.8337327, t1 = 0.5333333, a = 15


```{r}
all_results_400 |>
  ggplot(aes(x = h0, y = t1)) +
  geom_point(aes(color = algorithm)) +
  geom_hline(yintercept = 0.53, color = "red") +
  geom_vline(xintercept = 0.15, color = "red") +
  facet_wrap(~algorithm)
```

The bounds are different: lma can estimate h0 up to 0.5...

```{r}
all_results_400 |>
  filter(h0 == 0.35) |>
  summarise(h0_hit_upperbound = n())

316/750
```


```{r}
all_results_400 |>
  ggplot(aes(x = a, y = h1)) +
  geom_point(aes(color = algorithm)) +
  geom_hline(yintercept = 0.83, color = "red") +
  geom_vline(xintercept = 15, color = "red") +
  facet_wrap(~algorithm)
```

h1 is consistently underestimated, except for some instances when a is underestimated (interesting). Estimation for a are kind of all over the place

```{r}
lowest_sse_df_400 <- all_results_400 |>
  group_by(dataset_id) |>
  mutate(lowest_sse = min(best_sse)) |>
  filter(best_sse == lowest_sse) |>
  select(dataset_id, algorithm, lowest_sse)

lowest_sse_df_400 |>
  group_by(algorithm) |>
  summarise(count = n())

```


# double-sigmoidal data

```{r}
# f_base
f0 <- function(x, B1, M1, B2, L) {
  1 / ((1 + exp(-B1 * (x - M1))) * (1 + exp(B2 * (x - (M1 + L)))))
}

# piecewise double sigmoid

f2_h0 <- function (x, A2, Ka, B1, M1, B2, L, const, argument, h0){
  fBasics::Heaviside(x - argument) * (f0(x, B1, M1, B2, L) * ((Ka - A2 * Ka)/(const)) + A2 * Ka) +
    (1 - fBasics::Heaviside(x - argument)) * (f0(x, B1, M1, B2, L) * ((Ka - h0)/(const)) + h0) }


```


```{r}
# generate data
generate_piecewise_data <- function(x_vals = seq(0, 300, by = 20), reps = 5, dispersion = 100,
    B1 = 0.04, B2 = 0.03, M1 = 110, L = 130, h1 = 480, h2 = 240, h0 = 0) {
  A2 <- h2 / h1
  Ka <- h1

  # set.seed(101)
  
  # find x-value corresponding to max of f0 (which is t*)
  x_dense <- seq(min(x_vals), max(x_vals), length.out = 1000)
  f0_vals <- f0(x_dense, B1, M1, B2, L)
  argument <- x_dense[which.max(f0_vals)]

  # const = f0
  const <- f0(argument, B1, M1, B2, L) # argument because max(f0) is at t*

  # add reps to x and y
  x_repeated <- rep(x_vals, each = reps)
  y_true <- f2_h0(x_repeated, A2, Ka, B1, M1, B2, L, const, argument, h0)
  y_noisy <- rnorm(length(y_true), mean = y_true, sd = dispersion)

  # make df
  df <- data.frame(
    time = x_repeated,
    intensity = y_noisy)
    
  norm_df <- df |>
  mutate(time = ((time - min(time))/(max(time)- min(time))),
         intensity = ((intensity - min(intensity))/(max(intensity)- min(intensity))))
  
  return(norm_df)
  
}
```

Objective Function

```{r}
objective_SSE_double <- function(x, data) {
  h0 <- x[1]
  h1 <- x[2]
  h2 <- x[3]
  a1 <- x[4]
  a2 <- x[5]
  t1 <- x[6]
  t2 <- x[7]
  
  sse <- data |>
    mutate(
      preds = h0 +
              (h1 - h0) / (1 + exp(-a1 * (time - t1))) +
              (h1 - h2) / (1 + exp(-a2 * (time - t2))),
      diff = (intensity - preds)^2
    ) |>
    summarize(sse = sum(diff)) |>
    pull()
  
  return(sse)
}

```


```{r}

seeds <- 1:250 

# generate the data with given seeds
double_sig_data <- map(seeds, function(s) {
  set.seed(s)
  generate_piecewise_data()
})

# define bounds
bounds <- matrix(c(-0.10,  0.30, 0.00,   0.01,   0.01,  -0.52, -0.48,
                   0.30,  0.97, 1.00, 180.00, 180.00,   1.15,  1.79), nrow = 2, byrow = TRUE)


# helper function to run everything smoothly in parallel
run_woa_parallel <- function(data, i, algo_func, algo_name) {
  out <- withSeed({
    algo_func(data)
  }, seed = seeds[i])  # ensures all random numbers inside algo_func are reproducible

  out$dataset_id <- i
  out$seed <- seeds[i]
  out$algorithm <- algo_name
  out
}

# og WOA
double_og_df <- future_map2_dfr(
  double_sig_data, seq_along(double_sig_data),
  ~ run_woa_parallel(.x, .y, run_WOA, "WOA"),
  seed = TRUE  # ensures parallel-safe independent RNG streams
)

# chaotic WOA
double_chaotic_df <- future_map2_dfr(
  double_sig_data, seq_along(double_sig_data),
  ~ run_woa_parallel(.x, .y, run_WOA_chaotic, "WOA_chaotic"),
  seed = TRUE
)

# quasirandom WOA
double_quasi_df <- future_map2_dfr(
  double_sig_data, seq_along(double_sig_data),
  ~ run_woa_parallel(.x, .y, run_WOA_quasirandom, "WOA_quasirandom"),
  seed = TRUE
)

# combine
all_results_double <- bind_rows(double_og_df, double_chaotic_df, double_quasi_df)

```
This is taking forever to run, so let me try to run just one:

```{r}
# define data
set.seed(47)
double_data <- generate_piecewise_data()

# objective function is SSE
objective_SSE_double <- function(x, data) {
  h0 <- x[1]
  h1 <- x[2]
  h2 <- x[3]
  a1 <- x[4]
  a2 <- x[5]
  t1 <- x[6]
  t2 <- x[7]
  
  sse <- data |>
    mutate(
      preds = h0 +
              (h1 - h0) / (1 + exp(-a1 * (time - t1))) +
              (h1 - h2) / (1 + exp(-a2 * (time - t2))),
      diff = (intensity - preds)^2
    ) |>
    summarize(sse = sum(diff)) |>
    pull()
  
  return(sse)
}

# upper and lower bounds
bounds <- matrix(c(-0.10,  0.30, 0.00,   0.01,   0.01,  -0.52, -0.48, # this messes up midpoint distance
                   0.30,  0.97, 1.00, 180.00, 180.00,   1.15,  1.79), nrow = 2, byrow = TRUE)


tic()
# run WOA
sse_result <- WOA_quasirandom(
  FUN = function(x) objective_SSE_double(x, double_data),
  optimType = "MIN",
  numVar = ncol(bounds),
  numPopulation = 40, # default 40
  maxIter = 500, # default 500
  rangeVar = bounds)

toc()

best_sse <- objective_SSE_double(sse_result, double_data)

print(sse_result)
print(best_sse)
```


```{r}
# define double_sig_formula for knowing the order for plotting
# doublesigmoidalFitFormula_h0 <- function(x,
#                                       finalAsymptoteIntensityRatio,
#                                       maximum,
#                                       slope1Param,
#                                       midPoint1Param,
#                                       slope2Param,
#                                       midPointDistanceParam,
#                                       h0)

double_data |>
  ggplot(aes(x = time, y = intensity)) +
  geom_point() +
  geom_line(aes(x = time, y = sicegar:::doublesigmoidalFitFormula_h0(time, 0.2865116/0.7174965, 0.7174965, 49.1378929, 0.3663755, 128.4211427, 1.2151188-0.3663755, 0.2297220)), color = "red")
```



# Levy flight

Even though most of this work was done in a separate doc, I want it hear so i can rbind to all_results.

First define the functions:

```{r}
engine_WOA_levy_at_end <- function(FUN, optimType, maxIter, lowerBound, upperBound, whale) {
  whaleFitness <- calcFitness(FUN, optimType, whale)
  index <- order(whaleFitness)
  whaleFitness <- sort(whaleFitness)
  whale <- whale[index, ]
  bestPos <- whale[1, ]
  FbestPos <- whaleFitness[1]
  curve <- c()
  progressbar <- txtProgressBar(min = 0, max = maxIter, style = 3)
  for (t in 1:maxIter) {
    a <- 2 - t * ((2)/maxIter)
    a2 <- -1 + t * ((-1)/maxIter)
    for (i in 1:nrow(whale)) {
      r1 <- runif(1)
      r2 <- runif(1)
      A <- 2 * a * r1 - a
      C <- 2 * r2
      b <- 1
      l <- (a2 - 1) * runif(1) + 1
      p <- runif(1)
      for (j in 1:ncol(whale)) {
        if (p < 0.5) {
          if (abs(A) >= 1) {
            rand.index <- floor(nrow(whale) * runif(1) + 
              1)
            x.rand <- whale[rand.index, ]
            D.x.rand <- abs(C * x.rand[j] - whale[i, 
              j])
            whale[i, j] <- x.rand[j] - A * D.x.rand
          }
          else if (abs(A) < 1) {
            D.bestPos <- abs(C * bestPos[j] - whale[i, 
              j])
            whale[i, j] <- bestPos[j] - A * D.bestPos
          }
        }
        else if (p >= 0.5) {
          distance <- abs(bestPos[j] - whale[i, j])
          whale[i, j] <- distance * exp(b * l) * cos(l * 
            2 * pi) + bestPos[j]
        }
      }
      whale[i, ] <- checkBound(whale[i, ], lowerBound, 
        upperBound)
      fitness <- optimType * FUN(whale[i, ])
      if (fitness < FbestPos) {
        FbestPos <- fitness
        bestPos <- whale[i, ]
      }
    }
    
# this is the levy part
    
scalar <- 0.01  # step-size scaling which we could tune...

for (i in 1:nrow(whale)) {
  # generate levy vector for each whale
  s <- levy_step(n = ncol(whale), beta = 1.5)
  
  # move whale by one levy step
  whale[i, ] <- whale[i, ] + scalar * s
  
  # check bounds
  whale[i, ] <- checkBound(whale[i, ], lowerBound, upperBound)
  
  # recalc fitness
  fitness <- optimType * FUN(whale[i, ])
  if (fitness < FbestPos) {
    FbestPos <- fitness
    bestPos <- whale[i, ]
  }
}
    # end of levy part
    curve[t] <- FbestPos
    setTxtProgressBar(progressbar, t)
  }
  close(progressbar)
  curve <- curve * optimType
  return(bestPos)
}
```



```{r}
WOA_levy_at_end <- function(FUN, optimType = "MIN", numVar, numPopulation = 40, 
  maxIter = 500, rangeVar) 
{
  dimension <- ncol(rangeVar)
  lowerBound <- rangeVar[1, ]
  upperBound <- rangeVar[2, ]
  if (dimension == 1) {
    dimension <- numVar
  }
  if (optimType == "MAX") 
    optimType <- -1
  else optimType <- 1
  whale <- generateRandom(numPopulation, dimension, lowerBound, 
    upperBound)
  bestPos <- engine_WOA_levy_at_end(FUN, optimType, maxIter, lowerBound, 
    upperBound, whale)
  return(bestPos)
}
```

Test it:

```{r}
bounds <- matrix(c(-0.1, 0.3, -0.52, 0.01,
                 0.35, 0.97, 1, 180), nrow = 2, ncol = 4, byrow = TRUE)

# run WOA_quasirandom
run_WOA_levy_at_end <- function(data) {
  # running WOA_quasirandom
  sse_result <- WOA_levy_at_end(
    FUN = function(x) objective_SSE(x, data),
    optimType = "MIN",
    numVar = ncol(bounds),
    numPopulation = 40,
    maxIter = 500,
    rangeVar = bounds
  )
  
  # compute best SSE at that solution
  best_sse <- objective_SSE(sse_result, data)
  
  # return both
  list(
    h0 = sse_result[1],
    h1 = sse_result[2],
    t1 = sse_result[3],
    a  = sse_result[4],
    best_sse = best_sse
  )
}
tic()
run_WOA_levy_at_end(norm_sig_data)
toc()
```


Then I want to run on the same 250 datasets...

```{r}
library(furrr)
library(R.utils)

tic()
WOA_leavy_at_end_df <- future_map2_dfr(
  generated_data, seq_along(generated_data),
  ~ run_woa_parallel(.x, .y, run_WOA_levy_at_end, "WOA_levy_at_end"),
  seed = TRUE  # ensures parallel-safe independent RNG streams
)
toc()
```




