---
title: "hybrid_alg"
format: html
---

```{r}
library(tidyverse)
library(tictoc)
```

```{r}
source("WOA_functions.R")
```


The idea for the hybrid function is that the transition is triggered by a diversity score with WOA - when it shrinks to a certain threshold (5% diversity) the alg switches to lma. diversity is calculated by finding the median whale in each dimension, then for each dimension d, it computes the mean absolute deviation from the median across all N whales, then averages those D values. If this doesn't happen within a certain number of iterations, then the switch to lma will automatically happen, (default is 200). When the diversity has shrunk to 5% of it's initial value, lma is run from the best k whales as long as they are sufficiently far apart. k is capped at 10, but is a hyperparameter set by the user between 1 and 10. In terms of chosing which whales should be used as starting points for lma, min_dist = 0.05 means a whale must be at least 5% of the search space diameter away from all existing whales. I'm not yet sure if this is the right value...


# woa_lma_hybrid

```{r}
# ============================================================
# WOA-LMA Hybrid Optimizer
# ============================================================
# Packages: minpack.lm, metaheuristicOpt (for generateRandom
#           and checkBound helpers)
# install.packages(c("minpack.lm", "metaheuristicOpt"))
#
# DESIGN
# ------
# Phase 1 — WOA global search
#   Runs the standard WOA loop iteration by iteration.
#   At each iteration, population diversity is computed using the
#   median-based formula from Nadimi-Shahraki et al. (2023):
#     Diversity(t) = (1/D) * sum_d [ (1/N) * sum_i |median{x_d(t)} - x_id(t)| ]
#   i.e. for each dimension d, compute the mean absolute deviation
#   from the median across all N whales, then average over all D dims.
#   Transition to Phase 2 is triggered when:
#     (a) diversity drops below `div_threshold`, OR
#     (b) the iteration cap `max_iter_woa` is reached.
#
# Phase 2 — LMA local refinement
#   The top `n_lma` whales by fitness are selected as LMA seeds.
#   A minimum distance filter (`min_dist`) removes seeds that
#   are too close to each other, keeping the most diverse set.
#   LMA is run from each surviving seed via nls.lm().
#   The best converged LMA result is returned as the final answer.
#
# INPUTS
# ------
#   FUN           : objective function f(params) -> scalar
#   residual_fn   : residual function for LMA, r(params) -> vector
#                   such that sum(r^2) = FUN(params).
#                   If NULL, a generic single-residual wrapper is
#                   used (less accurate Jacobian, but still works).
#   numVar        : number of variables
#   rangeVar      : 2 x numVar matrix, row 1 = lower, row 2 = upper
#   numPopulation : WOA population size         (default 40)
#   max_iter_woa  : hard iteration cap for WOA  (default 200)
#   div_threshold : diversity trigger (normalised, 0-1)
#                   set relative to initial diversity (default 0.05)
#   n_lma         : max number of LMA seeds     (default 5, max 10)
#   min_dist      : minimum Euclidean distance between seeds
#                   as fraction of search space diameter (default 0.05)
#   lm_control    : nls.lm.control() list passed to each LMA run
#   verbose       : print progress messages     (default TRUE)
#
# OUTPUT
# ------
# A list with:
#   $par            : best parameter vector found
#   $value          : objective value at $par
#   $woa_iters      : number of WOA iterations run
#   $trigger        : "diversity" or "max_iter"
#   $lma_results    : list of all individual LMA outputs
#   $lma_best_idx   : index of the winning LMA seed
#   $woa_history    : data.frame of per-iteration WOA best fitness
#                     and diversity (useful for plotting)
#   $seeds_used     : matrix of LMA starting positions after filtering
# ============================================================

library(minpack.lm)
library(metaheuristicOpt)   # for generateRandom() and checkBound()

# ------------------------------------------------------------
# MAIN HYBRID FUNCTION
# ------------------------------------------------------------
woa_lma_hybrid <- function(
    FUN,
    residual_fn   = NULL,
    numVar,
    rangeVar,
    numPopulation = 40,
    max_iter_woa  = 200,
    div_threshold = 0.05,      # fraction of initial diversity
    n_lma         = 5,
    min_dist      = 0.05,      # fraction of search space diameter
    lm_control    = nls.lm.control(maxiter = 500, ftol = 1e-12,
                                    ptol = 1e-12),
    verbose       = TRUE
) {

  # ---- input checks ----------------------------------------
  n_lma <- min(max(n_lma, 1), 10)
  stopifnot(nrow(rangeVar) == 2, ncol(rangeVar) == numVar)

  lowerBound <- rangeVar[1, ]
  upperBound <- rangeVar[2, ]

  # search space diameter — used to scale min_dist
  space_diam <- sqrt(sum((upperBound - lowerBound)^2))
  abs_min_dist <- min_dist * space_diam

  # ---- generic residual wrapper if none provided -----------
  # Single residual: r = sqrt(f + eps). LMA still converges but
  # the Jacobian is less informative than a natural formulation.
  if (is.null(residual_fn)) {
    if (verbose) message(
      "[hybrid] No residual_fn supplied — using sqrt(FUN) wrapper.\n",
      "         For best LMA performance, supply a natural residual."
    )
    residual_fn <- function(params) sqrt(FUN(params) + 1e-10)
  }

  # ---- helper: population diversity -----------------------
  # Median-based diversity (Nadimi-Shahraki et al., 2023):
  #   Diversity(t) = (1/D) * sum_d [ (1/N) * sum_i |median{x_d} - x_id| ]
  # For each dimension d, compute the mean absolute deviation of all
  # whale positions from the median of that dimension, then average
  # those values across all D dimensions.
  compute_diversity <- function(whale_mat) {
    D       <- ncol(whale_mat)
    dim_mad <- vapply(seq_len(D), function(d) {
      col_d <- whale_mat[, d]
      mean(abs(col_d - median(col_d)))
    }, numeric(1))
    mean(dim_mad)
  }

  # ---- initialise WOA population --------------------------
  if (verbose) message("[WOA] Initialising population (n = ", numPopulation, ") ...")

  whale        <- generateRandom(numPopulation, numVar, lowerBound, upperBound)
  whaleFitness <- apply(whale, 1, FUN)

  idx          <- order(whaleFitness)
  whaleFitness <- whaleFitness[idx]
  whale        <- whale[idx, , drop = FALSE]

  bestPos      <- whale[1, ]
  FbestPos     <- whaleFitness[1]

  # record initial diversity as baseline for threshold
  init_diversity <- compute_diversity(whale)
  trigger_level  <- div_threshold * init_diversity

  if (verbose) message(
    "[WOA] Initial best fitness : ", round(FbestPos, 6),
    "\n[WOA] Initial diversity    : ", round(init_diversity, 6),
    "\n[WOA] Diversity trigger at : ", round(trigger_level, 6)
  )

  # storage for history
  woa_history <- data.frame(
    iter      = integer(),
    best_fit  = numeric(),
    diversity = numeric()
  )

  trigger      <- "max_iter"
  woa_iters    <- 0L

  # ---- WOA main loop --------------------------------------
  for (t in seq_len(max_iter_woa)) {

    a  <- 2 - t * (2 / max_iter_woa)     # decreases 2 -> 0
    a2 <- -1 + t * (-1 / max_iter_woa)   # decreases -1 -> -2

    for (i in seq_len(nrow(whale))) {

      r1 <- runif(1); r2 <- runif(1)
      A  <- 2 * a * r1 - a
      C  <- 2 * r2
      b  <- 1
      l  <- (a2 - 1) * runif(1) + 1
      p  <- runif(1)

      for (j in seq_len(ncol(whale))) {
        if (p < 0.5) {
          if (abs(A) >= 1) {
            # exploration: move toward random whale
            rand_idx   <- floor(nrow(whale) * runif(1) + 1)
            x_rand     <- whale[rand_idx, ]
            D_rand     <- abs(C * x_rand[j] - whale[i, j])
            whale[i, j] <- x_rand[j] - A * D_rand
          } else {
            # exploitation: shrinking encircling
            D_best     <- abs(C * bestPos[j] - whale[i, j])
            whale[i, j] <- bestPos[j] - A * D_best
          }
        } else {
          # spiral update
          dist_to_best <- abs(bestPos[j] - whale[i, j])
          whale[i, j]  <- dist_to_best * exp(b * l) * cos(l * 2 * pi) + bestPos[j]
        }
      }

      # enforce bounds
      whale[i, ] <- checkBound(whale[i, ], lowerBound, upperBound)

      # evaluate and update best
      fit <- FUN(whale[i, ])
      if (fit < FbestPos) {
        FbestPos <- fit
        bestPos  <- whale[i, ]
      }
    }

    woa_iters  <- t
    diversity  <- compute_diversity(whale)

    woa_history <- rbind(woa_history, data.frame(
      iter      = t,
      best_fit  = FbestPos,
      diversity = diversity
    ))

    if (verbose && t %% 25 == 0) {
      message("[WOA] iter ", t, " | best = ", round(FbestPos, 6),
              " | diversity = ", round(diversity, 6))
    }

    # --- diversity transition check -------------------------
    if (diversity < trigger_level) {
      trigger <- "diversity"
      if (verbose) message(
        "[WOA] Diversity trigger fired at iteration ", t,
        " (diversity = ", round(diversity, 6), " < ",
        round(trigger_level, 6), ")"
      )
      break
    }
  }

  if (trigger == "max_iter" && verbose) {
    message("[WOA] Max iteration cap reached (", max_iter_woa, " iters).")
  }

  # ---- select LMA seeds ------------------------------------
  # Sort all whales by fitness and take top n_lma
  final_fitness <- apply(whale, 1, FUN)
  top_idx       <- order(final_fitness)[seq_len(min(n_lma, nrow(whale)))]
  candidates    <- whale[top_idx, , drop = FALSE]

  if (verbose) message(
    "\n[LMA] Selecting seeds from top ", nrow(candidates),
    " whales ..."
  )

  # ---- minimum distance filter -----------------------------
  # Greedily keep seeds that are >= abs_min_dist apart.
  # Always keep the best (first) candidate.
  selected <- candidates[1, , drop = FALSE]

  if (nrow(candidates) > 1) {
    for (k in 2:nrow(candidates)) {
      candidate_k <- candidates[k, ]
      dists_to_selected <- apply(selected, 1, function(s) {
        sqrt(sum((candidate_k - s)^2))
      })
      if (all(dists_to_selected >= abs_min_dist)) {
        selected <- rbind(selected, candidate_k)
      } else {
        if (verbose) message(
          "[LMA] Seed ", k, " skipped (too close to an existing seed, ",
          "min dist = ", round(min(dists_to_selected), 6), ")"
        )
      }
    }
  }

  rownames(selected) <- NULL

  if (verbose) message(
    "[LMA] Running LMA from ", nrow(selected), " seed(s) ..."
  )

  # ---- run LMA from each seed ------------------------------
  lma_results  <- vector("list", nrow(selected))
  lma_values   <- rep(Inf, nrow(selected))

  for (k in seq_len(nrow(selected))) {
    seed_k <- as.numeric(selected[k, ])

    if (verbose) message(
      "[LMA] Seed ", k, ": starting at (",
      paste(round(seed_k, 4), collapse = ", "), ")"
    )

    result_k <- tryCatch(
      nls.lm(par = seed_k, fn = residual_fn, control = lm_control),
      error = function(e) {
        message("[LMA] Seed ", k, " errored: ", conditionMessage(e))
        NULL
      }
    )

    lma_results[[k]] <- result_k

    if (!is.null(result_k)) {
      val_k        <- FUN(result_k$par)
      lma_values[k] <- val_k

      if (verbose) message(
        "[LMA] Seed ", k, ": converged to (",
        paste(round(result_k$par, 5), collapse = ", "),
        ") | f = ", round(val_k, 8),
        " | info = ", result_k$info,
        if (result_k$info %in% c(1,2,3,4)) " [OK]" else " [!check convergence]"
      )
    }
  }

  # ---- select best LMA result ------------------------------
  best_lma_idx <- which.min(lma_values)
  best_result  <- lma_results[[best_lma_idx]]

  if (is.null(best_result)) {
    # all LMA runs failed — fall back to WOA best
    warning("[hybrid] All LMA runs failed. Returning WOA best position.")
    final_par <- bestPos
    final_val <- FbestPos
  } else {
    final_par <- best_result$par
    final_val <- lma_values[best_lma_idx]
  }

  if (verbose) message(
    "\n[hybrid] DONE",
    "\n  WOA trigger  : ", trigger, " (at iter ", woa_iters, ")",
    "\n  LMA winner   : seed ", best_lma_idx,
    "\n  Final params : (", paste(round(final_par, 6), collapse = ", "), ")",
    "\n  Final f      : ", round(final_val, 10)
  )

  return(list(
    par          = final_par,
    value        = final_val,
    woa_iters    = woa_iters,
    trigger      = trigger,
    lma_results  = lma_results,
    lma_best_idx = best_lma_idx,
    woa_history  = woa_history,
    seeds_used   = selected
  ))
}

```

```{r}
# ============================================================
# EXAMPLE USAGE — Rastrigin function
# ============================================================
# Rastrigin: f(x,y) = 20 + x^2 - 10cos(2pi*x) + y^2 - 10cos(2pi*y)
# Global minimum at (0, 0), f = 0.
# No natural least-squares formulation, so we use the sqrt wrapper.

rastrigin <- function(params) {
  x <- params[1]; y <- params[2]
  20 + x^2 - 10*cos(2*pi*x) + y^2 - 10*cos(2*pi*y)
}

set.seed(42)

result_rastrigin <- woa_lma_hybrid(
  FUN           = rastrigin,
  residual_fn   = NULL,          # will use sqrt wrapper automatically
  numVar        = 2,
  rangeVar      = matrix(c(-5, 5, -5, 5), nrow = 2),
  numPopulation = 40,
  max_iter_woa  = 200,
  div_threshold = 0.05,
  n_lma         = 5,
  min_dist      = 0.05,
  verbose       = TRUE
)


# ============================================================
# EXAMPLE USAGE — Rosenbrock function
# ============================================================
# Rosenbrock: f(x,y) = (1-x)^2 + 100*(y - x^2)^2
# Global minimum at (1, 1), f = 0.
# Natural residual formulation available — much better for LMA.

rosenbrock <- function(params) {
  x <- params[1]; y <- params[2]
  (1 - x)^2 + 100*(y - x^2)^2
}

rosenbrock_residual <- function(params) {
  x <- params[1]; y <- params[2]
  c(1 - x, 10*(y - x^2))    # sum of squares = rosenbrock(params)
}

set.seed(42)

result_rosenbrock <- woa_lma_hybrid(
  FUN           = rosenbrock,
  residual_fn   = rosenbrock_residual,
  numVar        = 2,
  rangeVar      = matrix(c(-2, 2, -0.5, 3), nrow = 2),
  numPopulation = 40,
  max_iter_woa  = 200,
  div_threshold = 0.05,
  n_lma         = 5,
  min_dist      = 0.05,
  verbose       = TRUE
)
```

first, the rastrigin doesn't have a natural least squares formulation which is why the first one is funky, however the execution of the alg worked correctly.

if we use the rosenbrock as an example, we see that it worked perfectly (though this is an easy problem). we note that woa found the right answer after 50 iterations but the diversity didn't collapse until 170. then (as was true in both problems) it only ran lma from one whale since the candidates were so close. Either this won't be an issue on harder problems or, i will have to toggle the min_dist = 0.05.


# default woa

```{r}
run_woa_default_rastrigin <- function(seed, n_dim = 5, numPopulation = 200,
                                       maxIter = 300) {
  set.seed(seed)
  
  bounds <- matrix(
    c(rep(-5.12, n_dim), rep(5.12, n_dim)),
    nrow = 2, byrow = TRUE
  )
  
  result <- WOA(
    FUN           = rastrigin,
    optimType     = "MIN",
    numVar        = n_dim,
    numPopulation = numPopulation,
    maxIter       = maxIter,
    rangeVar      = bounds
  )
  
  tibble(
    seed      = seed,
    value     = rastrigin(result),
    algorithm = "WOA_default"
  )
}

run_woa_default_schwefel <- function(seed, n_dim = 5, numPopulation = 200,
                                      maxIter = 300) {
  set.seed(seed)
  
  bounds <- matrix(
    c(rep(-500, n_dim), rep(500, n_dim)),
    nrow = 2, byrow = TRUE
  )
  
  result <- WOA(
    FUN           = schwefel,
    optimType     = "MIN",
    numVar        = n_dim,
    numPopulation = numPopulation,
    maxIter       = maxIter,
    rangeVar      = bounds
  )
  
  tibble(
    seed      = seed,
    value     = schwefel(result),
    algorithm = "WOA_default"
  )
}
```


# woa 75%

This func only runs until woa has reached 25% of its og diversity

```{r}
run_woa_early_stop_rastrigin <- function(seed, n_dim = 5, numPopulation = 200,
                                          maxIter = 300, div_threshold = 0.25) {
  set.seed(seed)
  
  bounds <- matrix(
    c(rep(-5.12, n_dim), rep(5.12, n_dim)),
    nrow = 2, byrow = TRUE
  )
  
  lowerBound <- bounds[1, ]
  upperBound <- bounds[2, ]
  
  compute_diversity <- function(whale_mat) {
    D       <- ncol(whale_mat)
    dim_mad <- vapply(seq_len(D), function(d) {
      col_d <- whale_mat[, d]
      mean(abs(col_d - median(col_d)))
    }, numeric(1))
    mean(dim_mad)
  }
  
  # initialize population
  whale        <- generateRandom(numPopulation, n_dim, lowerBound, upperBound)
  whaleFitness <- apply(whale, 1, rastrigin)
  idx          <- order(whaleFitness)
  whaleFitness <- whaleFitness[idx]
  whale        <- whale[idx, , drop = FALSE]
  bestPos      <- whale[1, ]
  FbestPos     <- whaleFitness[1]
  
  init_diversity <- compute_diversity(whale)
  trigger_level  <- div_threshold * init_diversity
  
  # standard WOA loop (no normalization, no levy) with diversity stopping
  for (t in seq_len(maxIter)) {
    a  <- 2 - t * (2 / maxIter)
    a2 <- -1 + t * (-1 / maxIter)
    
    for (i in seq_len(nrow(whale))) {
      r1 <- runif(1); r2 <- runif(1)
      A  <- 2 * a * r1 - a
      C  <- 2 * r2
      b  <- 1
      l  <- (a2 - 1) * runif(1) + 1
      p  <- runif(1)
      
      for (j in seq_len(ncol(whale))) {
        if (p < 0.5) {
          if (abs(A) >= 1) {
            rand_idx    <- floor(nrow(whale) * runif(1) + 1)
            x_rand      <- whale[rand_idx, ]
            D_rand      <- abs(C * x_rand[j] - whale[i, j])
            whale[i, j] <- x_rand[j] - A * D_rand
          } else {
            D_best      <- abs(C * bestPos[j] - whale[i, j])
            whale[i, j] <- bestPos[j] - A * D_best
          }
        } else {
          dist_to_best <- abs(bestPos[j] - whale[i, j])
          whale[i, j]  <- dist_to_best * exp(b * l) * cos(l * 2 * pi) + bestPos[j]
        }
      }
      
      whale[i, ] <- checkBound(whale[i, ], lowerBound, upperBound)
      
      fit <- rastrigin(whale[i, ])
      if (fit < FbestPos) {
        FbestPos <- fit
        bestPos  <- whale[i, ]
      }
    }
    
    # check diversity trigger
    if (compute_diversity(whale) < trigger_level) break
  }
  
  tibble(
    seed      = seed,
    value     = FbestPos,
    algorithm = "WOA_early_stop"
  )
}
```



```{r}
run_woa_early_stop_schwefel <- function(seed, n_dim = 5, numPopulation = 200,
                                         maxIter = 300, div_threshold = 0.25) {
  set.seed(seed)
  
  bounds <- matrix(
    c(rep(-500, n_dim), rep(500, n_dim)),
    nrow = 2, byrow = TRUE
  )
  
  lowerBound <- bounds[1, ]
  upperBound <- bounds[2, ]
  
  compute_diversity <- function(whale_mat) {
    D       <- ncol(whale_mat)
    dim_mad <- vapply(seq_len(D), function(d) {
      col_d <- whale_mat[, d]
      mean(abs(col_d - median(col_d)))
    }, numeric(1))
    mean(dim_mad)
  }
  
  whale        <- generateRandom(numPopulation, n_dim, lowerBound, upperBound)
  whaleFitness <- apply(whale, 1, schwefel)
  idx          <- order(whaleFitness)
  whaleFitness <- whaleFitness[idx]
  whale        <- whale[idx, , drop = FALSE]
  bestPos      <- whale[1, ]
  FbestPos     <- whaleFitness[1]
  
  init_diversity <- compute_diversity(whale)
  trigger_level  <- div_threshold * init_diversity
  
  for (t in seq_len(maxIter)) {
    a  <- 2 - t * (2 / maxIter)
    a2 <- -1 + t * (-1 / maxIter)
    
    for (i in seq_len(nrow(whale))) {
      r1 <- runif(1); r2 <- runif(1)
      A  <- 2 * a * r1 - a
      C  <- 2 * r2
      b  <- 1
      l  <- (a2 - 1) * runif(1) + 1
      p  <- runif(1)
      
      for (j in seq_len(ncol(whale))) {
        if (p < 0.5) {
          if (abs(A) >= 1) {
            rand_idx    <- floor(nrow(whale) * runif(1) + 1)
            x_rand      <- whale[rand_idx, ]
            D_rand      <- abs(C * x_rand[j] - whale[i, j])
            whale[i, j] <- x_rand[j] - A * D_rand
          } else {
            D_best      <- abs(C * bestPos[j] - whale[i, j])
            whale[i, j] <- bestPos[j] - A * D_best
          }
        } else {
          dist_to_best <- abs(bestPos[j] - whale[i, j])
          whale[i, j]  <- dist_to_best * exp(b * l) * cos(l * 2 * pi) + bestPos[j]
        }
      }
      
      whale[i, ] <- checkBound(whale[i, ], lowerBound, upperBound)
      
      fit <- schwefel(whale[i, ])
      if (fit < FbestPos) {
        FbestPos <- fit
        bestPos  <- whale[i, ]
      }
    }
    
    if (compute_diversity(whale) < trigger_level) break
  }
  
  tibble(
    seed      = seed,
    value     = FbestPos,
    algorithm = "WOA_early_stop"
  )
}
```





# woa_levy_lma_hybrid

```{r}
woa_levy_lma_hybrid <- function(
    FUN,
    residual_fn   = NULL,
    numVar,
    rangeVar,
    numPopulation = 40,
    max_iter_woa  = 200,
    div_threshold = 0.05,
    n_lma         = 5,
    min_dist      = 0.05,
    beta          = 1.5,
    lm_control    = nls.lm.control(maxiter = 500, ftol = 1e-12,
                                    ptol = 1e-12),
    verbose       = TRUE
) {

  # ---- input checks ----------------------------------------
  n_lma <- min(max(n_lma, 1), 10)
  stopifnot(nrow(rangeVar) == 2, ncol(rangeVar) == numVar)

  lowerBound   <- rangeVar[1, ]
  upperBound   <- rangeVar[2, ]
  space_diam   <- sqrt(sum((upperBound - lowerBound)^2))
  abs_min_dist <- min_dist * space_diam

  # ---- generic residual wrapper if none provided -----------
  if (is.null(residual_fn)) {
    if (verbose) message(
      "[hybrid] No residual_fn supplied — using sqrt(FUN) wrapper.\n",
      "         For best LMA performance, supply a natural residual."
    )
    residual_fn <- function(params) sqrt(FUN(params) + 1e-10)
  }

  # ---- helper: population diversity ------------------------
  compute_diversity <- function(whale_mat) {
    D       <- ncol(whale_mat)
    dim_mad <- vapply(seq_len(D), function(d) {
      col_d <- whale_mat[, d]
      mean(abs(col_d - median(col_d)))
    }, numeric(1))
    mean(dim_mad)
  }

  # ---- initialise WOA population --------------------------
  if (verbose) message("[WOA] Initialising population (n = ", numPopulation, ") ...")

  whale        <- generateRandom(numPopulation, numVar, lowerBound, upperBound)
  whaleFitness <- apply(whale, 1, FUN)
  idx          <- order(whaleFitness)
  whaleFitness <- whaleFitness[idx]
  whale        <- whale[idx, , drop = FALSE]
  bestPos      <- whale[1, ]
  FbestPos     <- whaleFitness[1]

  init_diversity <- compute_diversity(whale)
  trigger_level  <- div_threshold * init_diversity

  if (verbose) message(
    "[WOA] Initial best fitness : ", round(FbestPos, 6),
    "\n[WOA] Initial diversity    : ", round(init_diversity, 6),
    "\n[WOA] Diversity trigger at : ", round(trigger_level, 6)
  )

  woa_history <- data.frame(
    iter      = integer(),
    best_fit  = numeric(),
    diversity = numeric()
  )

  trigger   <- "max_iter"
  woa_iters <- 0L

  # ---- WOA main loop (levy exploration + normalized steps) -
  for (t in seq_len(max_iter_woa)) {

    a  <- 2 - t * (2 / max_iter_woa)
    a2 <- -1 + t * (-1 / max_iter_woa)

    for (i in seq_len(nrow(whale))) {

      r1 <- runif(1); r2 <- runif(1)
      A  <- 2 * a * r1 - a
      C  <- 2 * r2
      b  <- 1
      l  <- (a2 - 1) * runif(1) + 1
      p  <- runif(1)

      # normalize current whale and best position for step calculation
      whale_i_norm     <- normalize_params(whale[i, ], rangeVar)
      bestPos_norm     <- normalize_params(bestPos, rangeVar)
      whale_i_norm_new <- whale_i_norm

      for (j in seq_len(ncol(whale))) {
        if (p < 0.5) {
          if (abs(A) >= 1) {
            # levy exploration: jump toward random whale
            # scaled by distance to that whale
            rand_idx    <- floor(nrow(whale) * runif(1) + 1)
            x_rand_norm <- normalize_params(whale[rand_idx, ], rangeVar)

            s <- levy_step(n = 1, beta = beta)
            whale_i_norm_new[j] <- x_rand_norm[j] + s * abs(x_rand_norm[j] - whale_i_norm[j])

          } else {
            # standard normalized exploitation (unchanged)
            D.bestPos           <- abs(C * bestPos_norm[j] - whale_i_norm[j])
            whale_i_norm_new[j] <- bestPos_norm[j] - A * D.bestPos
          }

        } else {
          # standard normalized spiral (unchanged)
          distance            <- abs(bestPos_norm[j] - whale_i_norm[j])
          whale_i_norm_new[j] <- distance * exp(b * l) * cos(l * 2 * pi) + bestPos_norm[j]
        }
      }

      # denormalize back to original space
      whale[i, ] <- denormalize_params(whale_i_norm_new, rangeVar)

      # check bounds in original space
      whale[i, ] <- checkBound(whale[i, ], lowerBound, upperBound)

      # evaluate fitness in original space
      fit <- FUN(whale[i, ])
      if (fit < FbestPos) {
        FbestPos <- fit
        bestPos  <- whale[i, ]
      }
    }

    woa_iters <- t
    diversity <- compute_diversity(whale)

    woa_history <- rbind(woa_history, data.frame(
      iter      = t,
      best_fit  = FbestPos,
      diversity = diversity
    ))

    if (verbose && t %% 25 == 0) {
      message("[WOA] iter ", t, " | best = ", round(FbestPos, 6),
              " | diversity = ", round(diversity, 6))
    }

    # --- diversity transition check -------------------------
    if (diversity < trigger_level) {
      trigger <- "diversity"
      if (verbose) message(
        "[WOA] Diversity trigger fired at iteration ", t,
        " (diversity = ", round(diversity, 6), " < ",
        round(trigger_level, 6), ")"
      )
      break
    }
  }

  if (trigger == "max_iter" && verbose)
    message("[WOA] Max iteration cap reached (", max_iter_woa, " iters).")

  # ---- select LMA seeds ------------------------------------
  final_fitness <- apply(whale, 1, FUN)
  top_idx       <- order(final_fitness)[seq_len(min(n_lma, nrow(whale)))]
  candidates    <- whale[top_idx, , drop = FALSE]

  if (verbose) message(
    "\n[LMA] Selecting seeds from top ", nrow(candidates), " whales ..."
  )

  # ---- minimum distance filter -----------------------------
  selected <- candidates[1, , drop = FALSE]

  if (nrow(candidates) > 1) {
    for (k in 2:nrow(candidates)) {
      candidate_k       <- candidates[k, ]
      dists_to_selected <- apply(selected, 1, function(s) sqrt(sum((candidate_k - s)^2)))
      if (all(dists_to_selected >= abs_min_dist)) {
        selected <- rbind(selected, candidate_k)
      } else {
        if (verbose) message(
          "[LMA] Seed ", k, " skipped (too close to an existing seed, ",
          "min dist = ", round(min(dists_to_selected), 6), ")"
        )
      }
    }
  }

  rownames(selected) <- NULL

  if (verbose) message("[LMA] Running LMA from ", nrow(selected), " seed(s) ...")

  # ---- run LMA from each seed ------------------------------
  lma_results <- vector("list", nrow(selected))
  lma_values  <- rep(Inf, nrow(selected))

  for (k in seq_len(nrow(selected))) {
    seed_k <- as.numeric(selected[k, ])

    if (verbose) message(
      "[LMA] Seed ", k, ": starting at (",
      paste(round(seed_k, 4), collapse = ", "), ")"
    )

    result_k <- tryCatch(
      nls.lm(par = seed_k, fn = residual_fn, control = lm_control),
      error = function(e) {
        message("[LMA] Seed ", k, " errored: ", conditionMessage(e))
        NULL
      }
    )

    lma_results[[k]] <- result_k

    if (!is.null(result_k)) {
      val_k         <- FUN(result_k$par)
      lma_values[k] <- val_k

      if (verbose) message(
        "[LMA] Seed ", k, ": converged to (",
        paste(round(result_k$par, 5), collapse = ", "),
        ") | f = ", round(val_k, 8),
        " | info = ", result_k$info,
        if (result_k$info %in% c(1,2,3,4)) " [OK]" else " [!check convergence]"
      )
    }
  }

  # ---- select best LMA result ------------------------------
  best_lma_idx <- which.min(lma_values)
  best_result  <- lma_results[[best_lma_idx]]

  if (is.null(best_result)) {
    warning("[hybrid] All LMA runs failed. Returning WOA best position.")
    final_par <- bestPos
    final_val <- FbestPos
  } else {
    final_par <- best_result$par
    final_val <- lma_values[best_lma_idx]
  }

  if (verbose) message(
    "\n[hybrid] DONE",
    "\n  WOA trigger  : ", trigger, " (at iter ", woa_iters, ")",
    "\n  LMA winner   : seed ", best_lma_idx,
    "\n  Final params : (", paste(round(final_par, 6), collapse = ", "), ")",
    "\n  Final f      : ", round(final_val, 10)
  )

  return(list(
    par          = final_par,
    value        = final_val,
    woa_iters    = woa_iters,
    trigger      = trigger,
    lma_results  = lma_results,
    lma_best_idx = best_lma_idx,
    woa_history  = woa_history,
    seeds_used   = selected
  ))
}
```



# Schweffel

```{r}

# schwefel

schwefel <- function(x) {
  n   <- length(x)
  val <- 418.9829 * n - sum(x * sin(sqrt(abs(x))))
  val
}


# schwefel_residual <- function(x) {
#   418.9829 - x * sin(sqrt(abs(x)))
# }

# schwefel_residual <- function(x) {
#   eps <- 1e-8
#   418.9829 - x * sin(sqrt(abs(x) + eps))
# }

schwefel_residual <- function(x) {
  sqrt(pmax(418.9829 - x * sin(sqrt(abs(x))), 0))
}

run_schwefel_benchmark <- function(n_dim = 5, seed = 42) {
  set.seed(seed)
  message("\n", strrep("=", 60))
  message("SCHWEFEL BENCHMARK  (", n_dim, "D)")
  message("Known global minimum: f = 0  at  x_i = 420.9687 for all i")
  message(strrep("=", 60))

  bounds_schwefel <- matrix(
    c(rep(-500, n_dim), rep(500, n_dim)),
    nrow = 2, byrow = TRUE
  )

  result <- woa_lma_hybrid(
    FUN         = schwefel,
    residual_fn = schwefel_residual,
    numVar      = n_dim,
    rangeVar    = bounds_schwefel,
    numPopulation = 200,
    max_iter_woa  = 300,
    div_threshold = 0.5,
    n_lma         = 10,
    min_dist      = 0.05,
    verbose       = TRUE
  )

  message("\n--- SCHWEFEL RESULTS ---")
  message("Best parameter values : ", paste(round(result$par, 4), collapse = ", "))
  message("Known optimum values  : ", paste(rep(420.9687, n_dim), collapse = ", "))
  message("Error per dimension   : ", paste(round(abs(result$par - 420.9687), 4), collapse = ", "))
  message("Final f(x)            : ", round(result$value, 8), "  (target = 0)")
  message("WOA iters used        : ", result$woa_iters, " (trigger: ", result$trigger, ")")

  invisible(result)
}

```


# Phase retrieval

```{r}
# ============================================================
# BENCHMARK 2: PHASE RETRIEVAL
# ============================================================
# Goal: recover a real signal x from its Fourier magnitude |F(x)|.
# The phase is lost — only |F(x)|^2 (power spectrum) is observed.
#
# Setup:
#   - True signal: a 1D Gaussian pulse on N = 16 points
#   - Observed data: |FFT(x_true)|^2  (magnitude squared, no phase)
#   - Parameters to recover: the N real signal values directly
#
# Objective (SSE in Fourier magnitude):
#   f(x) = sum_k ( |FFT(x)_k| - |FFT(x_true)_k| )^2
#
# Notes:
#   - The problem has trivial ambiguities (shift, reflection, conjugate flip)
#     so we look at the *magnitude spectrum* match, not parameter identity.
#   - Global minimum is 0 when magnitudes match exactly.
#   - LMA residuals: r_k = |FFT(x)_k| - |FFT(x_true)_k|
# ============================================================

make_phase_retrieval_problem <- function(N = 16, seed = 7) {
  set.seed(seed)

  # True signal: Gaussian pulse centred at N/2
  t_grid  <- seq_len(N)
  x_true  <- exp(-0.5 * ((t_grid - N/2) / (N/8))^2)

  # Observed Fourier magnitudes (what we know)
  fft_true     <- fft(x_true)
  obs_mag      <- Mod(fft_true)          # |F(x_true)|
  obs_mag_sq   <- obs_mag^2             # power spectrum (for reference)

  list(
    x_true   = x_true,
    obs_mag  = obs_mag,
    N        = N
  )
}

phase_retrieval_objective <- function(x, obs_mag) {
  fft_x   <- fft(x)
  rec_mag <- Mod(fft_x)
  sum((rec_mag - obs_mag)^2)
}

phase_retrieval_residuals <- function(x, obs_mag) {
  fft_x   <- fft(x)
  rec_mag <- Mod(fft_x)
  rec_mag - obs_mag   # vector of residuals, one per frequency bin
}

run_phase_retrieval_benchmark <- function(N = 16, seed = 42) {
  set.seed(seed)
  message("\n", strrep("=", 60))
  message("PHASE RETRIEVAL BENCHMARK  (N = ", N, " signal points)")
  message("Recover a 1D Gaussian signal from its Fourier magnitudes")
  message(strrep("=", 60))

  prob <- make_phase_retrieval_problem(N = N, seed = 7)

  # Bounds: signal values assumed in [-2, 2] (true signal is in [0,1])
  bounds_pr <- matrix(
    c(rep(-2, N), rep(2, N)),
    nrow = 2, byrow = TRUE
  )

  obj_fn  <- function(x) phase_retrieval_objective(x, prob$obs_mag)
  res_fn  <- function(x) phase_retrieval_residuals(x, prob$obs_mag)

  result <- woa_lma_hybrid(
    FUN           = obj_fn,
    residual_fn   = res_fn,
    numVar        = N,
    rangeVar      = bounds_pr,
    numPopulation = 60,
    max_iter_woa  = 300,
    div_threshold = 0.05,
    n_lma         = 5,
    min_dist      = 0.02,
    verbose       = TRUE
  )

  # Evaluate quality: compare recovered Fourier magnitudes to observed
  rec_mag  <- Mod(fft(result$par))
  mag_err  <- sqrt(mean((rec_mag - prob$obs_mag)^2))   # RMSE in magnitude

  message("\n--- PHASE RETRIEVAL RESULTS ---")
  message("Final f(x) (SSE in magnitude) : ", round(result$value, 8), "  (target = 0)")
  message("RMSE in Fourier magnitude     : ", round(mag_err, 8))
  message("WOA iters used                : ", result$woa_iters,
          " (trigger: ", result$trigger, ")")

  # Diagnostic: correlation between recovered and true signal
  # (accounts for the shift/reflection ambiguity)
  cor_val <- max(abs(cor(result$par, prob$x_true)),
                 abs(cor(rev(result$par), prob$x_true)))
  message("Max |correlation| with true signal (shift-invariant): ", round(cor_val, 6))

  invisible(list(result = result, problem = prob, mag_rmse = mag_err))
}

```

```{r}
# run both benchmarks

# Schwefel (5D) 
tic()
schwefel_result <- run_schwefel_benchmark(n_dim = 5, seed = 42)
toc()
# convergence plot

# plot(schwefel_result$woa_history$iter,
#      schwefel_result$woa_history$best_fit,
#      type = "l", col = "steelblue", lwd = 2,
#      xlab = "WOA Iteration", ylab = "Best Fitness",
#      main = "Schwefel (5D) — WOA convergence")
# abline(h = 0, col = "red", lty = 2)

```

```{r}

# Phase Retrieval (16-point signal) 
pr_result <- run_phase_retrieval_benchmark(N = 16, seed = 42)

# convergence plot

# plot(pr_result$result$woa_history$iter,
#      pr_result$result$woa_history$best_fit,
#      type = "l", col = "darkorange", lwd = 2,
#      xlab = "WOA Iteration", ylab = "Best Fitness (SSE)",
#      main = "Phase Retrieval (16pt) — WOA convergence")
# abline(h = 0, col = "red", lty = 2)
#
```

# levy woa proper

this is a proper levy with z score steps just to see if it works better

```{r}
# WOA with Normalized Steps + Levy Flight Exploration


levy_step <- function(n = 1, beta = 1.5) {
  sigma_u <- (gamma(1 + beta) * sin(pi * beta / 2) /
               (gamma((1 + beta) / 2) * beta * 2^((beta - 1) / 2)))^(1 / beta)
  u <- rnorm(n, mean = 0, sd = sigma_u)
  v <- rnorm(n, mean = 0, sd = 1)
  s <- u / abs(v)^(1 / beta)
  return(s)
}

engineWOA_normalized_levy <- function(FUN, optimType, maxIter, lowerBound, upperBound,
                                      whale, bounds_original, beta = 1.5) {
  
  whaleFitness <- calcFitness(FUN, optimType, whale)
  index        <- order(whaleFitness)
  whaleFitness <- sort(whaleFitness)
  whale        <- whale[index, ]
  bestPos      <- whale[1, ]
  FbestPos     <- whaleFitness[1]
  curve        <- c()
  progressbar  <- txtProgressBar(min = 0, max = maxIter, style = 3)
  
  for (t in 1:maxIter) {
    a  <- 2 - t * (2 / maxIter)
    a2 <- -1 + t * (-1 / maxIter)
    
    for (i in 1:nrow(whale)) {
      r1 <- runif(1)
      r2 <- runif(1)
      A  <- 2 * a * r1 - a
      C  <- 2 * r2
      b  <- 1
      l  <- (a2 - 1) * runif(1) + 1
      p  <- runif(1)
      
      # normalize current whale and best position for step calculation
      whale_i_norm     <- normalize_params(whale[i, ], bounds_original)
      bestPos_norm     <- normalize_params(bestPos, bounds_original)
      whale_i_norm_new <- whale_i_norm
      
      for (j in 1:ncol(whale)) {
        if (p < 0.5) {
          if (abs(A) >= 1) {
            # --- LEVY EXPLORATION ---
            # replaces standard random-whale step with levy-scaled jump
            # toward a random whale, scaled by distance to that whale.
            # occasional large levy steps help escape boundary traps.
            rand.index  <- floor(nrow(whale) * runif(1) + 1)
            x_rand_norm <- normalize_params(whale[rand.index, ], bounds_original)
            
            s <- levy_step(n = 1, beta = beta)
            whale_i_norm_new[j] <- x_rand_norm[j] + s * abs(x_rand_norm[j] - whale_i_norm[j])
            
          } else if (abs(A) < 1) {
            # standard normalized exploitation (unchanged)
            D.bestPos           <- abs(C * bestPos_norm[j] - whale_i_norm[j])
            whale_i_norm_new[j] <- bestPos_norm[j] - A * D.bestPos
          }
          
        } else if (p >= 0.5) {
          # standard normalized spiral (unchanged)
          distance            <- abs(bestPos_norm[j] - whale_i_norm[j])
          whale_i_norm_new[j] <- distance * exp(b * l) * cos(l * 2 * pi) + bestPos_norm[j]
        }
      }
      
      # denormalize back to original space
      whale[i, ] <- denormalize_params(whale_i_norm_new, bounds_original)
      
      # check bounds in original space
      whale[i, ] <- checkBound(whale[i, ], lowerBound, upperBound)
      
      # evaluate fitness in original space
      fitness <- optimType * FUN(whale[i, ])
      if (fitness < FbestPos) {
        FbestPos <- fitness
        bestPos  <- whale[i, ]
      }
    }
    
    curve[t] <- FbestPos
    setTxtProgressBar(progressbar, t)
  }
  
  close(progressbar)
  curve <- curve * optimType
  return(bestPos)
}

WOA_normalized_levy <- function(FUN, optimType = "MIN", numVar, numPopulation = 40,
                                 maxIter = 500, rangeVar, beta = 1.5) {
  
  dimension <- ncol(rangeVar)
  if (dimension == 1) dimension <- numVar
  if (optimType == "MAX") optimType <- -1 else optimType <- 1
  
  lowerBound <- rangeVar[1, ]
  upperBound <- rangeVar[2, ]
  
  whale <- generateRandom(numPopulation, dimension, lowerBound, upperBound)
  
  bestPos <- engineWOA_normalized_levy(FUN, optimType, maxIter, lowerBound,
                                       upperBound, whale, rangeVar, beta)
  return(bestPos)
}

# schwefel runner for comparison
run_woa_levy_schwefel <- function(seed, n_dim = 5, numPopulation = 40,
                                   maxIter = 500, beta = 1.5) {
  set.seed(seed)
  
  bounds <- matrix(
    c(rep(-500, n_dim), rep(500, n_dim)),
    nrow = 2, byrow = TRUE
  )
  
  result <- WOA_normalized_levy(
    FUN           = schwefel,
    optimType     = "MIN",
    numVar        = n_dim,
    numPopulation = numPopulation,
    maxIter       = maxIter,
    rangeVar      = bounds,
    beta          = beta
  )
  
  tibble(
    seed      = seed,
    value     = schwefel(result),
    algorithm = "WOA_levy"
  )
}

```




# Compare to regular woa and lma

```{r}
# reminder: global min = 0 at x_i = 420.9687 for all i
# bounds for all are [-500, 500]

schwefel <- function(x) {
  n <- length(x)
  418.9829 * n - sum(x * sin(sqrt(abs(x))))
}

# schwefel_residual <- function(x) sqrt(schwefel(x) + 1e-10)

# z score woa
run_woa_schwefel <- function(seed, n_dim = 5, numPopulation = 40, maxIter = 500) {
  set.seed(seed)
  
  bounds <- matrix(
    c(rep(-500, n_dim), rep(500, n_dim)),
    nrow = 2, byrow = TRUE
  )
  
  result <- WOA_normalized_steps(
    FUN           = schwefel,
    optimType     = "MIN",
    numVar        = n_dim,
    numPopulation = numPopulation,
    maxIter       = maxIter,
    rangeVar      = bounds
  )
  
  tibble(
    seed      = seed,
    value     = schwefel(result),
    algorithm = "WOA_normalized"
  )
}

# lma (w random starts like sicegar)
run_lma_schwefel <- function(seed, n_dim = 5,
                              max_success = 20, max_attempts = 500) {
  set.seed(seed)
  
  lowerbounds <- rep(-500, n_dim)
  upperbounds <- rep( 500, n_dim)
  
  fits       <- list()
  sse_values <- c()
  attempts   <- 0
  
  while (length(fits) < max_success && attempts < max_attempts) {
    attempts <- attempts + 1
    
    start <- runif(n_dim, min = lowerbounds, max = upperbounds)
    
    res <- tryCatch(
      nls.lm(
        par     = start,
        fn      = schwefel_residual,
        control = nls.lm.control(maxiter = 1000, ptol = 1e-8, ftol = 1e-8)
      ),
      error = function(e) NULL
    )
    
    if (!is.null(res)) {
      val <- schwefel(res$par)
      fits[[length(fits) + 1]] <- res$par
      sse_values <- c(sse_values, val)
    }
  }
  
  if (length(fits) == 0) {
    return(tibble(seed = seed, value = NA, algorithm = "LMA_random"))
  }
  
  best_idx <- which.min(sse_values)
  
  tibble(
    seed      = seed,
    value     = sse_values[best_idx],
    algorithm = "LMA_random"
  )
}

# woa-lma hybrid
run_hybrid_schwefel <- function(seed, n_dim = 5, numPopulation = 300,
                                 max_iter_woa = 500) {
  set.seed(seed)
  
  bounds <- matrix(
    c(rep(-500, n_dim), rep(500, n_dim)),
    nrow = 2, byrow = TRUE
  )
  
  result <- woa_lma_hybrid(
    FUN           = schwefel,
    residual_fn   = schwefel_residual,
    numVar        = n_dim,
    rangeVar      = bounds,
    numPopulation = numPopulation,
    max_iter_woa  = max_iter_woa,
    div_threshold = 0.25,
    n_lma         = 10,
    min_dist      = 0.05,
    verbose       = FALSE
  )
  
  tibble(
    seed      = seed,
    value     = result$value,
    algorithm = "WOA_LMA_hybrid"
  )
}

# woa-levy-lma hybrid
run_levy_hybrid_schwefel <- function(seed, n_dim = 5, numPopulation = 300,
                                 max_iter_woa = 500) {
  set.seed(seed)
  
  bounds <- matrix(
    c(rep(-500, n_dim), rep(500, n_dim)),
    nrow = 2, byrow = TRUE
  )
  
  result <- woa_levy_lma_hybrid(
    FUN           = schwefel,
    residual_fn   = schwefel_residual,
    numVar        = n_dim,
    rangeVar      = bounds,
    numPopulation = numPopulation,
    max_iter_woa  = max_iter_woa,
    div_threshold = 0.25,
    n_lma         = 10,
    min_dist      = 0.05,
    beta          = 1.5,
    verbose       = FALSE
  )
  
  tibble(
    seed      = seed,
    value     = result$value,
    algorithm = "WOA_Levy_LMA_hybrid"
  )
}


# running comparison
library(tictoc)

seeds <- 1

tic("WOA normalized")
woa_results <- purrr::map_dfr(seeds, run_woa_schwefel, numPopulation = 300, maxIter = 300)
toc()

tic("LMA random")
lma_results <- purrr::map_dfr(seeds, run_lma_schwefel, max_success = 20, max_attempts = 500)
toc()

tic("WOA-LMA hybrid")
hybrid_results <- purrr::map_dfr(seeds, run_hybrid_schwefel, numPopulation = 300, max_iter_woa = 300)
toc()

tic("WOA levy")
levy_results  <- purrr::map_dfr(seeds, run_woa_levy_schwefel, numPopulation = 300, maxIter = 300, beta = 1.5)
toc()

tic("WOA-levy-LMA hybrid")
levy_hybrid_results <- purrr::map_dfr(seeds, run_levy_hybrid_schwefel, numPopulation = 300, max_iter_woa = 300)
toc()

results <- bind_rows(woa_results, levy_results, lma_results, hybrid_results, levy_hybrid_results)


# summary
epsilon <- 0.01

results |>
  group_by(algorithm) |>
  summarise(
    mean_value   = mean(value,  na.rm = TRUE),
    median_value = median(value, na.rm = TRUE),
    min_value    = min(value,   na.rm = TRUE),
    max_value    = max(value,   na.rm = TRUE),
    n_failed     = sum(is.na(value)),
    n_epsilon    = sum(value < epsilon, na.rm = TRUE)
  )
```

First, it's worth noting that even with multiple random starts, some spots on the surface were not favorable (didn't play nicely with derivatives) and caused lma to fail. That said, LMA ran with 20 random starts and it's best value is 642 while WOA and the hybrid have values 0.00x.. We then note that the mean value and the minimum value for both WOA and the hybrid are very similar, however, the hybrid takes about half as much time to run (12.3 seconds vs 21.8 @ 200 search agents). Then it gets worse at 300 search agents??? 


mean val @ 700 population is 148
mean val @ 1000 population is 100
mean val @ 2500 population is 100

# rastrigin

Compare all algorithms: Rastrigin function (5D)
Global minimum: f = 0 at x_i = 0 for all i
Domain: x_i in [-5.12, 5.12]

```{r}
rastrigin <- function(x) {
  n <- length(x)
  10 * n + sum(x^2 - 10 * cos(2 * pi * x))
}

# rastrigin_residual <- function(theta) {
#   A <- 10
#   theta^2 - A * cos(2 * pi * theta) + A
# }

rastrigin_residual <- function(theta) {
  sqrt(pmax(theta^2 - 10 * cos(2 * pi * theta) + 10, 0))
}

# normalized woa for rastrigin

run_woa_rastrigin <- function(seed, n_dim = 5, numPopulation = 40, maxIter = 500) {
  set.seed(seed)
  
  bounds <- matrix(
    c(rep(-5.12, n_dim), rep(5.12, n_dim)),
    nrow = 2, byrow = TRUE
  )
  
  result <- WOA_normalized_steps(
    FUN           = rastrigin,
    optimType     = "MIN",
    numVar        = n_dim,
    numPopulation = numPopulation,
    maxIter       = maxIter,
    rangeVar      = bounds
  )
  
  tibble(
    seed      = seed,
    value     = rastrigin(result),
    algorithm = "WOA_normalized"
  )
}


# woa levy for rastrigin

run_woa_levy_rastrigin <- function(seed, n_dim = 5, numPopulation = 40,
                                    maxIter = 500, beta = 1.5) {
  set.seed(seed)
  
  bounds <- matrix(
    c(rep(-5.12, n_dim), rep(5.12, n_dim)),
    nrow = 2, byrow = TRUE
  )
  
  result <- WOA_normalized_levy(
    FUN           = rastrigin,
    optimType     = "MIN",
    numVar        = n_dim,
    numPopulation = numPopulation,
    maxIter       = maxIter,
    rangeVar      = bounds,
    beta          = beta
  )
  
  tibble(
    seed      = seed,
    value     = rastrigin(result),
    algorithm = "WOA_levy"
  )
}

# lma for rastrigin
run_lma_rastrigin <- function(seed, n_dim = 5,
                               max_success = 20, max_attempts = 500) {
  set.seed(seed)
  
  lowerbounds <- rep(-5.12, n_dim)
  upperbounds <- rep( 5.12, n_dim)
  
  fits       <- list()
  sse_values <- c()
  attempts   <- 0
  
  while (length(fits) < max_success && attempts < max_attempts) {
    attempts <- attempts + 1
    
    start <- runif(n_dim, min = lowerbounds, max = upperbounds)
    
    res <- tryCatch(
      nls.lm(
        par     = start,
        fn      = rastrigin_residual,
        control = nls.lm.control(maxiter = 1000, ptol = 1e-8, ftol = 1e-8)
      ),
      error = function(e) NULL
    )
    
    if (!is.null(res)) {
      val <- rastrigin(res$par)
      fits[[length(fits) + 1]] <- res$par
      sse_values <- c(sse_values, val)
    }
  }
  
  if (length(fits) == 0) {
    return(tibble(seed = seed, value = NA, algorithm = "LMA_random"))
  }
  
  best_idx <- which.min(sse_values)
  
  tibble(
    seed      = seed,
    value     = sse_values[best_idx],
    algorithm = "LMA_random"
  )
}


# woa-lma hybrid for rastrigin

run_hybrid_rastrigin <- function(seed, n_dim = 5, numPopulation = 200,
                                  max_iter_woa = 300) {
  set.seed(seed)
  
  bounds <- matrix(
    c(rep(-5.12, n_dim), rep(5.12, n_dim)),
    nrow = 2, byrow = TRUE
  )
  
  result <- woa_lma_hybrid(
    FUN           = rastrigin,
    residual_fn   = rastrigin_residual,
    numVar        = n_dim,
    rangeVar      = bounds,
    numPopulation = numPopulation,
    max_iter_woa  = max_iter_woa,
    div_threshold = 0.25,
    n_lma         = 10,
    min_dist      = 0.05,
    verbose       = FALSE
  )
  
  tibble(
    seed      = seed,
    value     = result$value,
    algorithm = "WOA_LMA_hybrid"
  )
}
```

```{r}
# add levy-lma hybrid runner
run_levy_hybrid_rastrigin <- function(seed, n_dim = 5, numPopulation = 200,
                                       max_iter_woa = 300) {
  set.seed(seed)
  
  bounds <- matrix(
    c(rep(-5.12, n_dim), rep(5.12, n_dim)),
    nrow = 2, byrow = TRUE
  )
  
  result <- woa_levy_lma_hybrid(
    FUN           = rastrigin,
    residual_fn   = rastrigin_residual,
    numVar        = n_dim,
    rangeVar      = bounds,
    numPopulation = numPopulation,
    max_iter_woa  = max_iter_woa,
    div_threshold = 0.25,
    n_lma         = 10,
    min_dist      = 0.05,
    beta          = 1.5,
    verbose       = FALSE
  )
  
  tibble(
    seed      = seed,
    value     = result$value,
    algorithm = "WOA_levy_LMA_hybrid"
  )
}

# running it all
seeds <- 21:40

tic("WOA normalized")
woa_results          <- purrr::map_dfr(seeds, run_woa_rastrigin,
                                        numPopulation = 350, maxIter = 300)
toc()

tic("WOA levy")
levy_results         <- purrr::map_dfr(seeds, run_woa_levy_rastrigin,
                                        numPopulation = 350, maxIter = 300, beta = 1.5)
toc()

tic("LMA random")
lma_results          <- purrr::map_dfr(seeds, run_lma_rastrigin,
                                        max_success = 20, max_attempts = 500)
toc()

tic("WOA-LMA hybrid")
hybrid_results       <- purrr::map_dfr(seeds, run_hybrid_rastrigin,
                                        numPopulation = 350, max_iter_woa = 300)
toc()

tic("WOA-levy-LMA hybrid")
levy_hybrid_results  <- purrr::map_dfr(seeds, run_levy_hybrid_rastrigin,
                                        numPopulation = 350, max_iter_woa = 300)
toc()

results <- bind_rows(woa_results, levy_results, lma_results,
                     hybrid_results, levy_hybrid_results)

# summary
epsilon <- 0.01

results |>
  group_by(algorithm) |>
  summarise(
    mean_value   = mean(value,   na.rm = TRUE),
    median_value = median(value, na.rm = TRUE),
    min_value    = min(value,    na.rm = TRUE),
    max_value    = max(value,    na.rm = TRUE),
    n_failed     = sum(is.na(value)),
    n_epsilon    = sum(value < epsilon, na.rm = TRUE)
  )
```




# dataframes

I want to make big dataframes to look at:
```{r}
# rastrigin

# running it all
seeds <- 1:50

tic("WOA normalized")
woa_norm_results_rast <- purrr::map_dfr(seeds, run_woa_rastrigin,
                                        numPopulation = 300, maxIter = 300)
toc()

tic("WOA levy")
levy_results_rast <- purrr::map_dfr(seeds, run_woa_levy_rastrigin,
                                        numPopulation = 300, maxIter = 300, beta = 1.5)
toc()

tic("LMA random")
lma_results_rast  <- purrr::map_dfr(seeds, run_lma_rastrigin,
                                        max_success = 20, max_attempts = 500)
toc()

tic("WOA-LMA hybrid")
hybrid_results_rast <- purrr::map_dfr(seeds, run_hybrid_rastrigin,
                                        numPopulation = 300, max_iter_woa = 300)
toc()

tic("WOA 75%")
early_stop_results_rast <- purrr::map_dfr(seeds, run_woa_early_stop_rastrigin,
                                        numPopulation = 300, maxIter = 300)
toc()

tic("WOA-levy-LMA hybrid")
levy_hybrid_results_rast  <- purrr::map_dfr(seeds, run_levy_hybrid_rastrigin,
                                        numPopulation = 300, max_iter_woa = 300)
toc()

tic("WOA")
woa_default_results_rast <- purrr::map_dfr(seeds, run_woa_default_rastrigin, numPopulation = 300, maxIter = 300)
toc()

results_rast <- bind_rows(woa_norm_results_rast, levy_results_rast, lma_results_rast,
                     hybrid_results_rast, early_stop_results_rast, levy_hybrid_results_rast, woa_default_results_rast)

# summary
epsilon <- 0.01

all_algs_rastrigin_df <- results_rast |>
  group_by(algorithm) |>
  summarise(
    mean_value   = mean(value,   na.rm = TRUE),
    median_value = median(value, na.rm = TRUE),
    min_value    = min(value,    na.rm = TRUE),
    max_value    = max(value,    na.rm = TRUE),
    n_failed     = sum(is.na(value)),
    n_epsilon    = sum(value < epsilon, na.rm = TRUE))

# schwefel
tic("WOA normalized")
woa_norm_results_schwef <- purrr::map_dfr(seeds, run_woa_schwefel, numPopulation = 300, maxIter = 300)
toc()

tic("LMA random")
lma_results_schwef <- purrr::map_dfr(seeds, run_lma_schwefel, max_success = 20, max_attempts = 500)
toc()

tic("WOA-LMA hybrid")
hybrid_results_schwef <- purrr::map_dfr(seeds, run_hybrid_schwefel, numPopulation = 300, max_iter_woa = 300)
toc()

tic("WOA 75%")
early_stop_results_schwef <- purrr::map_dfr(seeds, run_woa_early_stop_schwefel,
                                        numPopulation = 300, maxIter = 300)
toc()

tic("WOA levy")
levy_results_schwef  <- purrr::map_dfr(seeds, run_woa_levy_schwefel, numPopulation = 300, maxIter = 300, beta = 1.5)
toc()

tic("WOA-levy-LMA hybrid")
levy_hybrid_results_schwef <- purrr::map_dfr(seeds, run_levy_hybrid_schwefel, numPopulation = 300, max_iter_woa = 300)
toc()

tic("WOA")
woa_default_results_schwef <- purrr::map_dfr(seeds, run_woa_default_schwefel, numPopulation = 300, maxIter = 300)
toc()

results_schwef <- bind_rows(woa_norm_results_schwef, levy_results_schwef, lma_results_schwef, hybrid_results_schwef, early_stop_results_schwef, levy_hybrid_results_schwef, woa_default_results_schwef)


# summary
epsilon <- 0.01

all_algs_schwefel_df <- results_schwef |>
  group_by(algorithm) |>
  summarise(
    mean_value   = mean(value,  na.rm = TRUE),
    median_value = median(value, na.rm = TRUE),
    min_value    = min(value,   na.rm = TRUE),
    max_value    = max(value,   na.rm = TRUE),
    n_failed     = sum(is.na(value)),
    n_epsilon    = sum(value < epsilon, na.rm = TRUE))
```

Run 2:
Rastrigin:
WOA normalized: 317.892 sec elapsed
WOA levy: 378.848 sec elapsed
LMA random: 0.575 sec elapsed
WOA-LMA hybrid: 27.398 sec elapsed
WOA-levy-LMA hybrid: 236.265 sec elapsed
WOA: 537.614 sec elapsed

Schwefel: - computer def fell asleep during some of these
WOA normalized: 337.571 sec elapsed
LMA random: 0.353 sec elapsed
WOA-LMA hybrid: 94.292 sec elapsed
WOA levy: 599.663 sec elapsed
WOA-levy-LMA hybrid: 5636.39 sec elapsed
WOA: 596.52 sec elapsed

Run 3:
Rastrigin:
WOA normalized: 361.323 sec elapsed
WOA levy: 349.316 sec elapsed
LMA random: 0.317 sec elapsed
WOA-LMA hybrid: 27.385 sec elapsed
WOA-levy-LMA hybrid: 230.005 sec elapsed
WOA: 185.525 sec elapsed

Schwefel:
WOA normalized: 297.028 sec elapsed
LMA random: 0.355 sec elapsed
WOA-LMA hybrid: 93.471 sec elapsed
WOA levy: 396.831 sec elapsed
WOA-levy-LMA hybrid: 143.09 sec elapsed
WOA: 226.48 sec elapsed


I want to check rastrigin results:

```{r}
# are the results actually zero?
woa_default_results_rast |> 
  summarise(exact_zeros = sum(value == 0))

# check vals
woa_default_results_rast |> pull(value) |> round(6)

```



```{r}
# check to see if there is an internal rastrigin
metaheuristicOpt:::rastrigin

```
There is!
Here's mine; they're the same:

function(x) {
  n <- length(x)
  10 * n + sum(x^2 - 10 * cos(2 * pi * x))
}


```{r}
# double check the result
set.seed(1)
bounds <- matrix(c(rep(-5.12, 5), rep(5.12, 5)), nrow = 2, byrow = TRUE)
result <- WOA(FUN = rastrigin, optimType = "MIN", numVar = 5,
              numPopulation = 300, maxIter = 300, rangeVar = bounds)
result
rastrigin(result)
```
It is indeed getting it right to 9 decimal points so really it is all zeros.


I think I need to try more dimensions


