---
title: "initial_WOA"
format: html
---

```{r}
library(tidyverse)
library(metaheuristicOpt)
library(tictoc)
```

### One Variable
Let's try to run the toy example of minimizing f(x) = (x-3)^2


```{r}
# define objective function
objective <- function(x) {
  (x[1] - 3)^2 # x[1] indexes it as the first variable
}

# define upper and lower bounds
bounds <- matrix(c(-50, 50,
                   -50, 50), nrow = 2, ncol = 2, byrow = FALSE)
# run WOA
result <- WOA(
  FUN = objective,
  optimType = "MIN",
  numVar = ncol(bounds),
  numPopulation = 5,
  maxIter = 50,
  rangeVar = bounds)

print(result)
```

For some reason it doesn't like when I run it with only one variable, so I added a dummy variable x2. Then it does a great job estimating x1 (which should be 3) and x2 is free.

### Two Variables
Let's actually try it with two variables:

```{r}
# define objective function
objective <- function(x) {
  (x[1] - 3)^2 + (x[2] - 2)^2 
}

# define upper and lower bounds
bounds <- matrix(c(-50, 50,
                   -50, 50), nrow = 2, ncol = 2, byrow = FALSE)
# run WOA
result <- WOA(
  FUN = objective,
  optimType = "MIN",
  numVar = ncol(bounds),
  numPopulation = 25,
  maxIter = 100,
  rangeVar = bounds)

print(result)
```

The answer should be (3,2). When numPopulation = 5 and maxIter = 50, I get (3.29, 1.2). When I increase numPopulation to 25 and maxIter to 100 I get the right answer.

### sicegar

#### nlsLM
Now, can we do it with sicegar data??

That means generating data, finding parameter estimates with LMA, and then finding corresponding SSE so that I have an idea of what the SSE should be, then trying to do it with WOA to see if we can get close.


```{r}
# generate data
set.seed(47)

# true values
h0 = 30
h1 = 750
t1 = 160
a = 0.05

reps = 5
disp = 250

sig_formula <- function(x, h0, h1, t1, a){
h0 + (h1 - h0) / (1 + exp(-a * (x - t1)))
}

x_vals <- rep(seq(0, 300, by = 20), times = reps)

true_params <- list(h0 = h0, h1 = h1, t1 = t1, a = a)

mean_vals <- rep(sig_formula(x_vals,
                        h0 = true_params$h0,
                        h1 = true_params$h1,
                        t1 = true_params$t1,
                        a = true_params$a))

data <- rnorm(n = length(mean_vals), mean = mean_vals, sd = rep(disp, length(mean_vals)))
sig_data <- data.frame(time = x_vals, intensity = data)

# also want a normalized version
norm_sig_data <- sig_data |>
  mutate(time = ((time - min(time))/(max(time)- min(time))),
         intensity = ((intensity - min(intensity))/(max(intensity)- min(intensity))))

```

Using `nlsLM`:

```{r}
# first define SSE function
sse_func <- function(h0, h1, t1, a, data) {
  sse <- data |>
    mutate(preds = h0 + (h1 - h0) / (1 + exp(-a * (x - t1))),
      diff = (y - preds)^2) |>
    summarize(sse = sum(diff)) |>
    pull()
  return(sse) # i'd rather this be a vector, but this only way i can get it to work down below
}
```


```{r}
# define bounds & starting values
startlist <- list(h0 = 0.13, h1 = 0.83, t1 = 0.53, a = 1) # changed starting values to be closer to true values

# had to make bigger bounds than we used with sicegar
lowerbounds <- c(h0 = -0.5, h1 = 0.0, t1 = -1, a = 0) 
upperbounds <- c(h0 = 0.5, h1 = 1.5, t1 = 1.5, a = 180) 

tic()
# run nlsLM
result <- minpack.lm::nlsLM(
  intensity ~ sig_formula(time, h0, h1, t1, a),
  data = norm_sig_data,
  start = startlist,
  control = list(maxiter = 1000, factor = 100, ptol = 1e-8, ftol = 1e-8),
  lower = lowerbounds,
  upper = upperbounds,
  trace = TRUE
)
toc()
# extract estimated parameters
params <- coef(result)
print(params)

# compute SSE
sse_val <- sse_func(
  h0 = params["h0"],            
  h1 = params["h1"],
  t1 = params["t1"],
  a  = params["a"],
  data = norm_sig_data |> rename(x = time, y = intensity) # match column names!
)

print(sse_val)
```

Maybe we should plot just to check it:

```{r}
norm_sig_data |>
  ggplot(aes(x = time, y = intensity)) +
  geom_point() +
  geom_line(aes(x = time, y = sig_formula(time, 0.1703829, 0.8272726, 0.5461193, 13.8626371)), color = "red")
```

```{r}
# checking normalization
# h1
h0_norm_func <- function(data, h0) {
  h0_norm <- (h0 - min(data$intensity))/(max(data$intensity) - min(data$intensity))
  return(h0_norm)
}


# h1
h1_norm_func <- function(data, h1) {
  h1_norm <- (h1 - min(data$intensity))/(max(data$intensity) - min(data$intensity))
  return(h1_norm)
}

# t1
t1_norm_func <- function(data, t1) {
  t1_norm <- (t1 - min(data$time))/(max(data$time) - min(data$time))
  return(t1_norm)
}

# a
a_norm_func <- function(data, a) {
  a_norm <- a * (max(data$time) - min(data$time))
  return(a_norm)
}

h0_norm_func(sig_data, h0)
h1_norm_func(sig_data, h1)
t1_norm_func(sig_data, t1)
a_norm_func(sig_data, a)
```

#### WOA

```{r}
# objective function is SSE
objective_SSE <- function(x, data) {
  h0 <- x[1]
  h1 <- x[2]
  t1 <- x[3]
  a  <- x[4]
  
  sse <- data |>
    mutate(preds = h0 + (h1 - h0)/(1 + exp(-a*(time - t1))),
           diff = (intensity - preds)^2) |>
    summarize(sse = sum(diff)) |>
    pull()
  
  return(sse)
}

# upper and lower bounds
bounds <- matrix(c(-0.1, 0.3, -0.52, 0.01,
                 0.3, 0.97, 1, 180), nrow = 2, ncol = 4, byrow = TRUE)
tic()
# run WOA
sse_result <- WOA(
  FUN = function(x) objective_SSE(x, norm_sig_data),
  optimType = "MIN",
  numVar = ncol(bounds),
  numPopulation = 40,
  maxIter = 500,
  rangeVar = bounds)

toc()

best_sse <- objective_SSE(sse_result, norm_sig_data)

print(sse_result)
print(best_sse)
```

Now we can plot:

```{r}
norm_sig_data |>
  ggplot(aes(x = time, y = intensity)) +
  geom_point() +
  geom_line(aes(x = time, y = sig_formula(time, 0.1658185, 0.8256365, 0.5442213, 14.0970407)), color = "red")
```

```{r}
sse_diff = best_sse - sse_val
sse_diff
```

