---
title: "levy_flight"
format: html
---

Here I'm trying to get a sense of what the Lévy flight distribution looks like. 

I'm going to use Mantegna's alogrithm which says:

levy($\lambda$) = $\frac{\mu}{|v|^\frac{1}{\lambda}}$ and perhaps we can think ab

This one is with $\beta$:

```{r}
# params
beta <- 1.5    
n <- 10000     

# defining sigma_u - we can use the gamma() in r
sigma_u <- (gamma(1 + beta) * sin(pi * beta / 2) /
               (gamma((1 + beta) / 2) * beta * 2^((beta - 1) / 2)))^(1 / beta)

# generate u and v
u <- rnorm(n, mean = 0, sd = sigma_u)
v <- rnorm(n, mean = 0, sd = 1)

# use to get step lengths
s <- u / (abs(v)^(1 / beta))

s <- data.frame(s)

# Visualize
s |>
  ggplot(aes(x = s)) +
  geom_histogram(bins = 200, fill = "violet", color = "black") +
  # coord_cartesian(xlim = c(-250, 250)) +
  labs(title = "Lévy flight Step Sizes",
       x = "Step length (s)", y = "Frequency") +
  theme_minimal()

quantile(s$s, probs = c(0, 0.01, 0.1, 0.5, 0.9, 0.99, 1))

```

This one is just with $\lambda$:

I could rewrite this easily to try different values of beta instead.

```{r}
# write function to run levy-flight with different lambda vals
levy_flight <- function(n = 10000, lambda = 1.5) {
  
  
  # same formula as above but lambda instead of beta
  sigma_u <- (gamma(1 + lambda) * sin(pi * lambda / 2) /
               (gamma((1 + lambda) / 2) * lambda * 2^((lambda - 1) / 2)))^(1 / lambda)
  
  # same u and v
  u <- rnorm(n, mean = 0, sd = sigma_u)
  v <- rnorm(n, mean = 0, sd = 1)
  
  # generate s
  s <- u / abs(v)^(1 / lambda)
  
  tibble(s = s, lambda = lambda)
}

# then run for different values of lambda
lambda_vals <- c(1.2, 1.5, 2.0) # for some reason 2.5 and 3 gave NA - must be with sin()

levy_data <- map_df(lambda_vals, ~levy_flight(n = 10000, lambda = .x))

# and plot
levy_data |>
  ggplot(aes(x = s)) +
  geom_histogram(bins = 500, fill = "violet", color = "black") +
  facet_wrap(~lambda) +
  coord_cartesian(xlim = c(-100, 100)) +
  labs(title = "Lévy Flight Step Sizes",
       x = "Step length (s)", y = "Frequency",
       subtitle = "λ values from (1, 2]") +
  theme_minimal()

```

```{r}

# write function to run levy-flight with different lambda vals
levy_flight_beta <- function(n = 10000, beta = 1.5) {
  
  
  # same formula as above but lambda instead of beta
  sigma_u <- (gamma(1 + beta) * sin(pi * beta / 2) /
               (gamma((1 + beta) / 2) * beta * 2^((beta - 1) / 2)))^(1 / beta)
  
  # same u and v
  u <- rnorm(n, mean = 0, sd = sigma_u)
  v <- rnorm(n, mean = 0, sd = 1)
  
  # generate s
  s <- u / abs(v)^(1 / beta)
  
  tibble(s = s, beta = beta)
}

# then run for different values of lambda
beta_vals <- c(1.2, 1.5, 1.8) # between 0 and 2

levy_data_beta <- map_df(beta_vals, ~levy_flight_beta(n = 10000, beta = .x))

# and plot
levy_data_beta |>
  ggplot(aes(x = s)) +
  geom_histogram(bins = 5000, fill = "violet", color = "black") +
  facet_wrap(~beta) +
  coord_cartesian(xlim = c(-100, 100)) +
  labs(title = "Lévy Flight Step Sizes",
       x = "Step length (s)", y = "Frequency",
       subtitle = "λ values from (1, 2]") +
  theme_minimal()
```

# combine with woa

```{r}
levy_step <- function(n = 10000, beta = 1.5) {
  
  
  # same formula as above but lambda instead of beta
  sigma_u <- (gamma(1 + beta) * sin(pi * beta / 2) /
               (gamma((1 + beta) / 2) * beta * 2^((beta - 1) / 2)))^(1 / beta)
  
  # same u and v
  u <- rnorm(n, mean = 0, sd = sigma_u)
  v <- rnorm(n, mean = 0, sd = 1)
  
  # generate s
  s <- u / abs(v)^(1 / beta)
  
  return(s)
}
```

New engineWOA: I am basically jittering every whale at the end by moving them by one levy step. I'm not sure if that is where it makes the most sense to do it. 

```{r}
engine_WOA_levy_at_end <- function(FUN, optimType, maxIter, lowerBound, upperBound, whale) {
  whaleFitness <- calcFitness(FUN, optimType, whale)
  index <- order(whaleFitness)
  whaleFitness <- sort(whaleFitness)
  whale <- whale[index, ]
  bestPos <- whale[1, ]
  FbestPos <- whaleFitness[1]
  curve <- c()
  progressbar <- txtProgressBar(min = 0, max = maxIter, style = 3)
  for (t in 1:maxIter) {
    a <- 2 - t * ((2)/maxIter)
    a2 <- -1 + t * ((-1)/maxIter)
    for (i in 1:nrow(whale)) {
      r1 <- runif(1)
      r2 <- runif(1)
      A <- 2 * a * r1 - a
      C <- 2 * r2
      b <- 1
      l <- (a2 - 1) * runif(1) + 1
      p <- runif(1)
      for (j in 1:ncol(whale)) {
        if (p < 0.5) {
          if (abs(A) >= 1) {
            rand.index <- floor(nrow(whale) * runif(1) + 
              1)
            x.rand <- whale[rand.index, ]
            D.x.rand <- abs(C * x.rand[j] - whale[i, 
              j])
            whale[i, j] <- x.rand[j] - A * D.x.rand
          }
          else if (abs(A) < 1) {
            D.bestPos <- abs(C * bestPos[j] - whale[i, 
              j])
            whale[i, j] <- bestPos[j] - A * D.bestPos
          }
        }
        else if (p >= 0.5) {
          distance <- abs(bestPos[j] - whale[i, j])
          whale[i, j] <- distance * exp(b * l) * cos(l * 
            2 * pi) + bestPos[j]
        }
      }
      whale[i, ] <- checkBound(whale[i, ], lowerBound, 
        upperBound)
      fitness <- optimType * FUN(whale[i, ])
      if (fitness < FbestPos) {
        FbestPos <- fitness
        bestPos <- whale[i, ]
      }
    }
    
# this is the levy part
    
scalar <- 0.01  # step-size scaling which we could tune...

for (i in 1:nrow(whale)) {
  # generate levy vector for each whale
  s <- levy_step(n = ncol(whale), beta = 1.5)
  
  # move whale by one levy step
  whale[i, ] <- whale[i, ] + scalar * s
  
  # check bounds
  whale[i, ] <- checkBound(whale[i, ], lowerBound, upperBound)
  
  # recalc fitness
  fitness <- optimType * FUN(whale[i, ])
  if (fitness < FbestPos) {
    FbestPos <- fitness
    bestPos <- whale[i, ]
  }
}
    # end of levy part
    curve[t] <- FbestPos
    setTxtProgressBar(progressbar, t)
  }
  close(progressbar)
  curve <- curve * optimType
  return(bestPos)
}
```


```{r}
WOA_levy_at_end <- function(FUN, optimType = "MIN", numVar, numPopulation = 40, 
  maxIter = 500, rangeVar) 
{
  dimension <- ncol(rangeVar)
  lowerBound <- rangeVar[1, ]
  upperBound <- rangeVar[2, ]
  if (dimension == 1) {
    dimension <- numVar
  }
  if (optimType == "MAX") 
    optimType <- -1
  else optimType <- 1
  whale <- generateRandom(numPopulation, dimension, lowerBound, 
    upperBound)
  bestPos <- engine_WOA_levy_at_end(FUN, optimType, maxIter, lowerBound, 
    upperBound, whale)
  return(bestPos)
}
```


```{r}
# define data
set.seed(47)
double_data <- generate_piecewise_data()

# objective function is SSE
objective_SSE_double <- function(x, data) {
  h0 <- x[1]
  h1 <- x[2]
  h2 <- x[3]
  a1 <- x[4]
  a2 <- x[5]
  t1 <- x[6]
  t2 <- x[7]
  
  sse <- data |>
    mutate(
      preds = h0 +
              (h1 - h0) / (1 + exp(-a1 * (time - t1))) +
              (h1 - h2) / (1 + exp(-a2 * (time - t2))),
      diff = (intensity - preds)^2
    ) |>
    summarize(sse = sum(diff)) |>
    pull()
  
  return(sse)
}

# upper and lower bounds
bounds <- matrix(c(-0.10,  0.30, 0.00,   0.01,   0.01,  -0.52, -0.48, # this messes up midpoint distance
                   0.30,  0.97, 1.00, 180.00, 180.00,   1.15,  1.79), nrow = 2, byrow = TRUE)


tic()
# run WOA
sse_result <- WOA_levy_at_end(
  FUN = function(x) objective_SSE_double(x, double_data),
  optimType = "MIN",
  numVar = ncol(bounds),
  numPopulation = 40, # default 40
  maxIter = 500, # default 500
  rangeVar = bounds)

toc()

best_sse <- objective_SSE_double(sse_result, double_data)

print(sse_result)
print(best_sse)
```











