---
title: "normalized_WOA"
format: html
---

---
title: "z_score_woa"
format: html
---

```{r}
# load libraries
library(tidyverse)
library(metaheuristicOpt)
library(minpack.lm)
library(furrr)
library(R.utils)
library(tictoc)
library(randtoolbox)


# source functions
source("WOA_functions.R")

```


Here I want to reprogram WOA to take steps using z-score's so that the crazy mismatch in scale between a and the other parameters does not impact the functions ability to converge to the minimum.

So,

# z score woa

```{r}
# define bounds
bounds <- matrix(c(-0.10,  0.30, 0.00,   0.01,   0.01,  -0.52, -0.48,
                   0.30,  0.97, 1.00, 180.00, 180.00,   1.15,  1.79), 
                 nrow = 2, byrow = TRUE)

# normalization functions
normalize_params <- function(params, bounds) {
  lower <- bounds[1, ]
  upper <- bounds[2, ]
  range <- upper - lower
  (params - lower) / range  # maps to [0, 1]
}

denormalize_params <- function(norm_params, bounds) {
  lower <- bounds[1, ]
  upper <- bounds[2, ]
  range <- upper - lower
  norm_params * range + lower  # maps back to original scale
}
```

```{r}
# Modified engineWOA - stays in original space, normalizes only for steps
engineWOA_normalized_steps <- function(FUN, optimType, maxIter, lowerBound, upperBound, 
                                       whale, bounds_original) {
  
  whaleFitness <- calcFitness(FUN, optimType, whale)
  index <- order(whaleFitness)
  whaleFitness <- sort(whaleFitness)
  whale <- whale[index, ]
  bestPos <- whale[1, ]
  FbestPos <- whaleFitness[1]
  curve <- c()
  progressbar <- txtProgressBar(min = 0, max = maxIter, style = 3)
  
  for (t in 1:maxIter) {
    a <- 2 - t * ((2)/maxIter)
    a2 <- -1 + t * ((-1)/maxIter)
    
    for (i in 1:nrow(whale)) {
      r1 <- runif(1)
      r2 <- runif(1)
      A <- 2 * a * r1 - a
      C <- 2 * r2
      b <- 1
      l <- (a2 - 1) * runif(1) + 1
      p <- runif(1)
      
      # NORMALIZE current whale and best position for step calculation
      whale_i_norm <- normalize_params(whale[i, ], bounds_original)
      bestPos_norm <- normalize_params(bestPos, bounds_original)
      
      # Create normalized copy for updates
      whale_i_norm_new <- whale_i_norm
      
      for (j in 1:ncol(whale)) {
        if (p < 0.5) {
          if (abs(A) >= 1) {
            rand.index <- floor(nrow(whale) * runif(1) + 1)
            # NORMALIZE random whale for distance calculation
            x_rand_norm <- normalize_params(whale[rand.index, ], bounds_original)
            
            D.x.rand <- abs(C * x_rand_norm[j] - whale_i_norm[j])
            whale_i_norm_new[j] <- x_rand_norm[j] - A * D.x.rand
          }
          else if (abs(A) < 1) {
            D.bestPos <- abs(C * bestPos_norm[j] - whale_i_norm[j])
            whale_i_norm_new[j] <- bestPos_norm[j] - A * D.bestPos
          }
        }
        else if (p >= 0.5) {
          distance <- abs(bestPos_norm[j] - whale_i_norm[j])
          whale_i_norm_new[j] <- distance * exp(b * l) * cos(l * 2 * pi) + bestPos_norm[j]
        }
      }
      
      # DENORMALIZE back to original space after taking normalized step
      whale[i, ] <- denormalize_params(whale_i_norm_new, bounds_original)
      
      # Check bounds in original space
      whale[i, ] <- checkBound(whale[i, ], lowerBound, upperBound)
      
      # Evaluate fitness in original space
      fitness <- optimType * FUN(whale[i, ])
      if (fitness < FbestPos) {
        FbestPos <- fitness
        bestPos <- whale[i, ]
      }
    }
    curve[t] <- FbestPos
    setTxtProgressBar(progressbar, t)
  }
  close(progressbar)
  curve <- curve * optimType
  
  # Return best position (already in original scale)
  return(bestPos)
}

# Modified WOA wrapper - works entirely in original space
WOA_normalized_steps <- function(FUN, optimType = "MIN", numVar, numPopulation = 40, 
                                 maxIter = 500, rangeVar) {
  
  dimension <- ncol(rangeVar)
  
  if (dimension == 1) {
    dimension <- numVar
  }
  
  if (optimType == "MAX") 
    optimType <- -1
  else 
    optimType <- 1
  
  # Original bounds (not normalized)
  lowerBound <- rangeVar[1, ]
  upperBound <- rangeVar[2, ]
  
  # Generate random whales in ORIGINAL parameter space
  whale <- generateRandom(numPopulation, dimension, lowerBound, upperBound)
  
  # Run optimization (stays in original space throughout)
  bestPos <- engineWOA_normalized_steps(FUN, optimType, maxIter, lowerBound, 
                                        upperBound, whale, rangeVar)
  
  return(bestPos)
}

# Usage
run_WOA_normalized_steps_double <- function(data) {
  result <- WOA_normalized_steps(
    FUN = function(x) objective_SSE_double(x, data),
    optimType = "MIN",
    numVar = 7,
    numPopulation = 40,
    maxIter = 2500,
    rangeVar = bounds
  )
  
  best_sse <- objective_SSE_double(result, data)
  
  list(
    h0 = result[1],
    h1 = result[2],
    h2 = result[3],
    a1 = result[4],
    a2 = result[5],
    t1 = result[6],
    t2 = result[7],
    best_sse = best_sse
  )
}
```


# test it

```{r}
# test it
# generate test data
set.seed(42)
test_data <- generate_piecewise_data(
  x_vals = seq(0, 300, by = 20), 
  reps = 5, 
  dispersion = 10,
  a1 = 0.04, a2 = 0.03, t1 = 110, t2 = 240, 
  h1 = 750, h2 = 400, h0 = 20
)



# Test it
result_norm <- run_WOA_normalized_steps_double(test_data)
print(result_norm)

result_og <- run_WOA_double(test_data)
print(result_og)
```



```{r, eval=FALSE}
# Run each 10 times to compare
set.seed(123)
results_norm <- replicate(10, run_WOA_normalized_steps_double(test_data)$best_sse)
set.seed(123)
results_og <- replicate(10, run_WOA_double(test_data)$best_sse)

cat("Normalized - Mean:", mean(results_norm), "SD:", sd(results_norm), "\n")
cat("Original - Mean:", mean(results_og), "SD:", sd(results_og), "\n")
```


Now that that's all set up, I want to see how the improved WOA (WOA_normalized) compares against LMA:

```{r}
# set up multicores
plan(multisession, workers = 5)  # i have 8 total

tic()
seeds <- 1:100

# generate the data with given seeds
double_sig_data <- map(seeds, function(s) {
  set.seed(s)
  generate_piecewise_data()
})

# helper function to run everything smoothly in parallel
run_woa_parallel <- function(data, i, algo_func, algo_name) {
  out <- withSeed({
    algo_func(data)
  }, seed = seeds[i])  # ensures all random numbers inside algo_func are reproducible

  out$dataset_id <- i
  out$seed <- seeds[i]
  out$algorithm <- algo_name
  out
}

# define bounds
bounds <- matrix(c(-0.10,  0.30, 0.00,   0.01,   0.01,  -0.52, -0.48,
                   0.30,  0.97, 1.00, 180.00, 180.00,   1.15,  1.79), nrow = 2, byrow = TRUE)

# woa w/ normalized step sizes
double_norm_df <- future_map2_dfr(
  double_sig_data, seq_along(double_sig_data),
  ~ run_woa_parallel(.x, .y, run_WOA_normalized_double, "WOA_norm"),
  seed = TRUE  # ensures parallel-safe independent RNG streams
)

# og woa
double_og_df <- future_map2_dfr(
  double_sig_data, seq_along(double_sig_data),
  ~ run_woa_parallel(.x, .y, run_WOA_double, "WOA"),
  seed = TRUE  # ensures parallel-safe independent RNG streams
)

# lma
lma_double_results_df <- map2_dfr(
  double_sig_data,  
  seeds,
  ~run_lma_fit_double(.x, dataset_id = which(seeds == .y), seed = .y)
)


norm_vs_lma_results_double <- bind_rows(double_norm_df, double_og_df, lma_double_results_df)
toc()
```


```{r}
norm_vs_lma_results_double |>
  ggplot(aes(x = dataset_id, y = best_sse)) +
  geom_point(aes(color = algorithm))
```


```{r}
norm_vs_lma_results_double |>
  filter(algorithm == "WOA") |>
  summarize(mean_sse = mean(best_sse))
```

Even with normalized steps, WOA is just bad at this problem. It is moving randomly around the valley and the spirals etc. I don't think match the downhill movement.





# verify for thesis

```{r}
# Modified engineWOA with dimension verification
engineWOA_normalized_steps <- function(FUN, optimType, maxIter, lowerBound, upperBound, 
                                       whale, bounds_original) {
  
  whaleFitness <- calcFitness(FUN, optimType, whale)
  index <- order(whaleFitness)
  whaleFitness <- sort(whaleFitness)
  whale <- whale[index, ]
  bestPos <- whale[1, ]
  FbestPos <- whaleFitness[1]
  curve <- c()
  progressbar <- txtProgressBar(min = 0, max = maxIter, style = 3)
  
  # Flag to print dimension info only once
  printed_dims <- FALSE
  
  for (t in 1:maxIter) {
    a <- 2 - t * ((2)/maxIter)
    a2 <- -1 + t * ((-1)/maxIter)
    
    # Print 'a' dimension check (only first iteration)
    if (t == 1) {
      cat("\n\n========== DIMENSION VERIFICATION (Iteration 1) ==========\n")
      cat(sprintf("'a' value: %.6f\n", a))
      cat(sprintf("'a' is scalar: %s\n", is.vector(a) && length(a) == 1))
      cat(sprintf("'a' dimensions: %s\n", paste(dim(a), collapse=" x ")))
      cat(sprintf("'a' length: %d\n", length(a)))
      cat(sprintf("'a2' value: %.6f\n", a2))
      cat("=========================================================\n\n")
    }
    
    for (i in 1:nrow(whale)) {
      r1 <- runif(1)
      r2 <- runif(1)
      A <- 2 * a * r1 - a
      C <- 2 * r2
      b <- 1
      l <- (a2 - 1) * runif(1) + 1
      p <- runif(1)
      
      # Print dimension checks for first whale in first iteration
      if (t == 1 && i == 1 && !printed_dims) {
        cat("\n========== SCALAR PARAMETER VERIFICATION ==========\n")
        cat(sprintf("Whale index: %d, Iteration: %d\n\n", i, t))
        
        # Check r1, r2
        cat("Random values (should be scalars):\n")
        cat(sprintf("  r1 = %.6f, length = %d, is.scalar = %s\n", 
                    r1, length(r1), length(r1) == 1))
        cat(sprintf("  r2 = %.6f, length = %d, is.scalar = %s\n", 
                    r2, length(r2), length(r2) == 1))
        
        # Check A, C
        cat("\nCoefficients (should be scalars):\n")
        cat(sprintf("  A = 2*a*r1 - a = 2*%.4f*%.4f - %.4f = %.6f\n", 
                    a, r1, a, A))
        cat(sprintf("  A length = %d, is.scalar = %s\n", length(A), length(A) == 1))
        cat(sprintf("  C = 2*r2 = 2*%.4f = %.6f\n", r2, C))
        cat(sprintf("  C length = %d, is.scalar = %s\n", length(C), length(C) == 1))
        
        # Check b, l, p
        cat("\nOther parameters (should be scalars):\n")
        cat(sprintf("  b = %d, length = %d\n", b, length(b)))
        cat(sprintf("  l = %.6f, length = %d\n", l, length(l)))
        cat(sprintf("  p = %.6f, length = %d\n", p, length(p)))
        
        cat("\n====================================================\n\n")
        printed_dims <- TRUE
      }
      
      # NORMALIZE current whale and best position for step calculation
      whale_i_norm <- normalize_params(whale[i, ], bounds_original)
      bestPos_norm <- normalize_params(bestPos, bounds_original)
      
      # Create normalized copy for updates
      whale_i_norm_new <- whale_i_norm
      
      # Track if A, r values change across dimensions
      if (t == 1 && i == 1) {
        cat("\n========== PER-DIMENSION UPDATE VERIFICATION ==========\n")
        cat(sprintf("Number of dimensions (parameters): %d\n", ncol(whale)))
        cat(sprintf("\nUsing SAME A = %.6f for ALL dimensions\n", A))
        cat(sprintf("Using SAME C = %.6f for ALL dimensions\n", C))
        cat(sprintf("Using SAME p = %.6f to determine update type\n\n", p))
        
        if (p < 0.5) {
          cat("Update type: Encircling prey or Search (p < 0.5)\n")
          if (abs(A) >= 1) {
            cat(sprintf("  |A| = %.4f >= 1: SEARCH (exploration)\n", abs(A)))
          } else {
            cat(sprintf("  |A| = %.4f < 1: ENCIRCLE (exploitation)\n", abs(A)))
          }
        } else {
          cat("Update type: Spiral (p >= 0.5)\n")
          cat(sprintf("  Using l = %.6f for spiral\n", l))
        }
        cat("\nDimension-by-dimension updates:\n")
      }
      
      for (j in 1:ncol(whale)) {
        # Store old value for comparison
        old_val <- whale_i_norm[j]
        
        if (p < 0.5) {
          if (abs(A) >= 1) {
            rand.index <- floor(nrow(whale) * runif(1) + 1)
            # NORMALIZE random whale for distance calculation
            x_rand_norm <- normalize_params(whale[rand.index, ], bounds_original)
            
            D.x.rand <- abs(C * x_rand_norm[j] - whale_i_norm[j])
            whale_i_norm_new[j] <- x_rand_norm[j] - A * D.x.rand
            
            # Print for first iteration, first whale
            if (t == 1 && i == 1) {
              cat(sprintf("  Dim %d: x_rand[%d]=%.4f, D=%.4f, new=%.4f (SAME A=%.4f used)\n", 
                          j, j, x_rand_norm[j], D.x.rand, whale_i_norm_new[j], A))
            }
          }
          else if (abs(A) < 1) {
            D.bestPos <- abs(C * bestPos_norm[j] - whale_i_norm[j])
            whale_i_norm_new[j] <- bestPos_norm[j] - A * D.bestPos
            
            # Print for first iteration, first whale
            if (t == 1 && i == 1) {
              cat(sprintf("  Dim %d: best[%d]=%.4f, D=%.4f, new=%.4f (SAME A=%.4f used)\n", 
                          j, j, bestPos_norm[j], D.bestPos, whale_i_norm_new[j], A))
            }
          }
        }
        else if (p >= 0.5) {
          distance <- abs(bestPos_norm[j] - whale_i_norm[j])
          whale_i_norm_new[j] <- distance * exp(b * l) * cos(l * 2 * pi) + bestPos_norm[j]
          
          # Print for first iteration, first whale
          if (t == 1 && i == 1) {
            cat(sprintf("  Dim %d: dist=%.4f, new=%.4f (SAME l=%.4f used)\n", 
                        j, distance, whale_i_norm_new[j], l))
          }
        }
      }
      
      if (t == 1 && i == 1) {
        cat("=======================================================\n\n")
        
        # Summary verification
        cat("========== SUMMARY VERIFICATION ==========\n")
        cat("✓ A, C, p, l are SCALARS (1x1)\n")
        cat("✓ SAME A value used across ALL dimensions\n")
        cat("✓ SAME C value used across ALL dimensions\n")
        cat("✓ SAME p value determines update type for ALL dimensions\n")
        cat("✓ SAME l value used for spiral across ALL dimensions\n")
        cat("✓ Each dimension j gets: new[j] = f(A, C, old[j], ...)\n")
        cat("✓ The update is VECTORIZED - same scalars, different positions\n")
        cat("==========================================\n\n")
      }
      
      # DENORMALIZE back to original space after taking normalized step
      whale[i, ] <- denormalize_params(whale_i_norm_new, bounds_original)
      
      # Check bounds in original space
      whale[i, ] <- checkBound(whale[i, ], lowerBound, upperBound)
      
      # Evaluate fitness in original space
      fitness <- optimType * FUN(whale[i, ])
      if (fitness < FbestPos) {
        FbestPos <- fitness
        bestPos <- whale[i, ]
      }
    }
    curve[t] <- FbestPos
    setTxtProgressBar(progressbar, t)
  }
  close(progressbar)
  curve <- curve * optimType
  
  # Return best position (already in original scale)
  return(bestPos)
}

# Modified WOA wrapper - works entirely in original space
WOA_normalized_steps <- function(FUN, optimType = "MIN", numVar, numPopulation = 40, 
                                 maxIter = 500, rangeVar) {
  
  dimension <- ncol(rangeVar)
  
  if (dimension == 1) {
    dimension <- numVar
  }
  
  if (optimType == "MAX") 
    optimType <- -1
  else 
    optimType <- 1
  
  # Original bounds (not normalized)
  lowerBound <- rangeVar[1, ]
  upperBound <- rangeVar[2, ]
  
  cat("\n========== WOA INITIALIZATION ==========\n")
  cat(sprintf("Number of variables (dimensions): %d\n", dimension))
  cat(sprintf("Population size: %d\n", numPopulation))
  cat(sprintf("Max iterations: %d\n", maxIter))
  cat(sprintf("Lower bounds: %s\n", paste(lowerBound, collapse=", ")))
  cat(sprintf("Upper bounds: %s\n", paste(upperBound, collapse=", ")))
  cat("========================================\n")
  
  # Generate random whales in ORIGINAL parameter space
  whale <- generateRandom(numPopulation, dimension, lowerBound, upperBound)
  
  cat(sprintf("\nWhale population matrix dimensions: %d x %d\n", nrow(whale), ncol(whale)))
  cat("(rows = whales, columns = parameters/dimensions)\n\n")
  
  # Run optimization (stays in original space throughout)
  bestPos <- engineWOA_normalized_steps(FUN, optimType, maxIter, lowerBound, 
                                        upperBound, whale, rangeVar)
  
  return(bestPos)
}

# Additional utility function to verify at any point in your code
verify_scalar <- function(var, var_name) {
  cat(sprintf("\nVerifying '%s':\n", var_name))
  cat(sprintf("  Value: %s\n", paste(var, collapse=", ")))
  cat(sprintf("  Length: %d\n", length(var)))
  cat(sprintf("  Is scalar (length=1): %s\n", length(var) == 1))
  cat(sprintf("  Class: %s\n", class(var)))
  if (!is.null(dim(var))) {
    cat(sprintf("  Dimensions: %s\n", paste(dim(var), collapse=" x ")))
  } else {
    cat("  Dimensions: NULL (not a matrix/array)\n")
  }
  cat(sprintf("  Is scalar: %s\n\n", is.vector(var) && length(var) == 1))
}

# Example usage demonstration
demo_dimension_check <- function() {
  cat("\n============ DEMONSTRATION ============\n\n")
  
  # Simulate the key calculations
  t <- 1
  maxIter <- 500
  a <- 2 - t * ((2)/maxIter)
  r1 <- runif(1)
  r2 <- runif(1)
  A <- 2 * a * r1 - a
  C <- 2 * r2
  
  verify_scalar(a, "a")
  verify_scalar(r1, "r1")
  verify_scalar(r2, "r2")
  verify_scalar(A, "A")
  verify_scalar(C, "C")
  
  cat("CONCLUSION:\n")
  cat("All parameters (a, r1, r2, A, C) are SCALARS (1x1).\n")
  cat("They are computed ONCE per whale per iteration.\n")
  cat("The SAME scalar values are used to update ALL dimensions.\n")
  cat("Each dimension j uses: new_pos[j] = f(A, old_pos[j], ...)\n")
  cat("where A is the SAME for all j.\n")
  cat("\n=======================================\n")
}

# Run the demonstration
demo_dimension_check()
```


