---
title: "running_everything"
format: html
---

This doc contains a mega chunk for running everything.

# Libraries

I am running WOA, WOA_chaotic, WOA_quasirandom, WOA_levy_at_end, LMA for sigmoidal data with dispersion 250, sigmoidal data with dispersion 400, and double sigmoidal data with dispersion 100.

```{r}
# load libraries
library(tidyverse)
library(metaheuristicOpt)
library(minpack.lm)
library(furrr)
library(R.utils)
library(tictoc)
library(randtoolbox)

```

# WOA helper functions

First we need to define this generateRandom(), calcFitness(), checkBound() functions, so I've just copied them from the og package.

```{r}
generateRandom <- function (numPopulation, dimension, lowerBound, upperBound) 
{
    result <- matrix()
    if (length(lowerBound) == 1) {
        result <- matrix(runif(numPopulation * dimension, lowerBound, 
            upperBound), nrow = numPopulation, ncol = dimension)
    }
    else {
        result <- matrix(nrow = numPopulation, ncol = dimension)
        for (i in 1:dimension) {
            result[, i] = runif(numPopulation, lowerBound[i], 
                upperBound[i])
        }
    }
    return(result)
}
```

```{r}
calcFitness <- function (FUN, optimType, popu) 
{
    fitness <- c()
    for (i in 1:nrow(popu)) {
        fitness[i] <- optimType * FUN(popu[i, ])
    }
    return(fitness)
}
```

```{r}
checkBound <- function (position, lowerBound, upperBound) 
{
    check1 <- position > upperBound
    check2 <- position < lowerBound
    result <- (position * (!(check1 + check2))) + upperBound * 
        check1 + lowerBound * check2
    return(result)
}
```

# SSE functions

## Sigmoidal

```{r}
objective_SSE <- function(x, data) {
  h0 <- x[1]
  h1 <- x[2]
  t1 <- x[3]
  a  <- x[4]
  
  sse <- data |>
    mutate(preds = h0 + (h1 - h0)/(1 + exp(-a*(time - t1))),
           diff = (intensity - preds)^2) |>
    summarize(sse = sum(diff)) |>
    pull()
  
  return(sse)
}
```

## Double sigmoidal

I'm going to recreate f0 and f2_h0 with the desired parameters instead of weird B1, M1 etc.

  A2 = h2/h1 
  Ka = h1
  B1 = a1
  M1 = t1
  B2 = a2
  L = t2 - t1
  h0 = h0

```{r}
# f_base
f0 <- function(x, a1, t1, a2, t2) {
  1 / ((1 + exp(-a1 * (x - t1))) * (1 + exp(a2 * (x - (t1 + (t2 - t1))))))
}

# piecewise double sigmoid
f2_h0 <- function (x, h0, h1, h2, a1, t1, a2, t2, const, argument){
  fBasics::Heaviside(x - argument) * (f0(x, a1, t1, a2, t2) * ((h1 - (h2/h1) * h1)/(const)) + (h2/h1) * h1) +
    (1 - fBasics::Heaviside(x - argument)) * (f0(x, a1, t1, a2, t2) * ((h1 - h0)/(const)) + h0) }


```

```{r}
double_sig_fit_formula <- function(time, h0, h1, h2, a1, a2, t1, t2){
  # A2 = h2/h1 # finalAsymptoteIntensityRatio
  # Ka = h1
  # B1 = a1
  # M1 = t1
  # B2 = a2
  # L = t2 - t1
  # h0 = h0
  
  # find x-value corresponding to max of f0 (which is t*)
  x_dense <- seq(min(time), max(time), length.out = 1000)
  f0_vals <- f0(x_dense, a1, t1, a2, t2)
  argument <- x_dense[which.max(f0_vals)]

  # const = f0
  const <- f0(argument, a1, t1, a2, t2) # argument because max(f0) is at t*
  
  f2_h0(time, h0, h1, h2, a1, t1, a2, t2, const, argument)
  
}
```

```{r}
sse_func_double <- function(h0, h1, h2, a1, a2, t1, t2, data) {
  sse <- data |>
    mutate(preds = double_sig_fit_formula(data$time, h0, h1, h2, a1, a2, t1, t2),
           diff = (data$intensity - preds)^2) |>
    summarize(sse = sum(diff)) |>
    pull()
  return(sse)
}
```

The only difference between `sse_func_double` and `objective_SSE_double` is that the former individual parameter estimates and the latter expects a vector.

I've also added checks to the function to make sure that h1 is positive, t2 is greater than t1, and a1 and a2 are non-zero.

```{r}
objective_SSE_double <- function(x, data) {
  h0 <- x[1]; h1 <- x[2]; h2 <- x[3]
  a1 <- x[4]; a2 <- x[5]; t1 <- x[6]; t2 <- x[7]
  
  # check for invalid parameters
  if (h1 <= 0 ||
    t2 <= t1 ||
    any(c(a1, a2) <= 0) ||
    h1 <= h0 ||
    h1 <= h2) {

  return(1e10) # big sse
}

  
  # try-catch in case double_sig_fit_formula still produces NA
  sse <- tryCatch({
    sse_func_double(h0, h1, h2, a1, a2, t1, t2, data)
  }, error = function(e) {
    1e10
  })
  
  if(is.na(sse) || is.nan(sse) || is.infinite(sse)) sse <- 1e10
  
  return(sse)
}
```

# WOA sigmoidal variants

## Chaotic

```{r}
engineWOA_chaotic <- function(FUN, optimType, maxIter, lowerBound, upperBound, whale) {
  whaleFitness <- calcFitness(FUN, optimType, whale)
  index <- order(whaleFitness)
  whaleFitness <- sort(whaleFitness)
  whale <- whale[index, ]
  bestPos <- whale[1, ]
  FbestPos <- whaleFitness[1]
  curve <- c()
  progressbar <- txtProgressBar(min = 0, max = maxIter, style = 3)

  # chaotic map
  chaotic_map <- function(x) 4 * x^3 - 3 * x
  
  # function to move range from (-1,1) to (0,1)
  to_unit <- function(x) (x + 1) / 2
  
  # starting value for r1, r2 - (-1,1) because that is domain of the map
  r1_val <- runif(1, -1, 1)
  r2_val <- runif(1, -1, 1)
 
  # back to regular function
  for (t in 1:maxIter) {
    a <- 2 - t * ((2) / maxIter)
    a2 <- -1 + t * ((-1) / maxIter)

    for (i in 1:nrow(whale)) {

      # update chaotic values
      r1_val <- chaotic_map(r1_val)
      r2_val <- chaotic_map(r2_val)
      

      # scale to [0,1]
      r1 <- to_unit(r1_val)
      r2 <- to_unit(r2_val)
      

      A <- 2 * a * r1 - a
      C <- 2 * r2
      b <- 1
      p  <- runif(1)
      l  <- (a2 - 1) * runif(1) + 1


      for (j in 1:ncol(whale)) {
        if (p < 0.5) {
          if (abs(A) >= 1) {
            rand.index <- floor(nrow(whale) * to_unit(chaotic_map(r1_val)) + 1)
            x.rand <- whale[rand.index, ]
            D.x.rand <- abs(C * x.rand[j] - whale[i, j])
            whale[i, j] <- x.rand[j] - A * D.x.rand
          } else {
            D.bestPos <- abs(C * bestPos[j] - whale[i, j])
            whale[i, j] <- bestPos[j] - A * D.bestPos
          }
        } else {
          distance <- abs(bestPos[j] - whale[i, j])
          whale[i, j] <- distance * exp(b * l) * cos(l * 2 * pi) + bestPos[j]
        }
      }

      whale[i, ] <- checkBound(whale[i, ], lowerBound, upperBound)
      fitness <- optimType * FUN(whale[i, ])
      if (fitness < FbestPos) {
        FbestPos <- fitness
        bestPos <- whale[i, ]
      }
    }
    curve[t] <- FbestPos
    setTxtProgressBar(progressbar, t)
  }
  close(progressbar)
  curve <- curve * optimType
  return(bestPos)
}

```

```{r}
WOA_chaotic <- function(FUN, optimType = "MIN", numVar, numPopulation = 40, 
  maxIter = 500, rangeVar) 
{
  dimension <- ncol(rangeVar)
  lowerBound <- rangeVar[1, ]
  upperBound <- rangeVar[2, ]
  if (dimension == 1) {
    dimension <- numVar
  }
  if (optimType == "MAX") 
    optimType <- -1
  else optimType <- 1
  whale <- generateRandom(numPopulation, dimension, lowerBound, 
    upperBound)
  bestPos <- engineWOA_chaotic(FUN, optimType, maxIter, lowerBound, 
    upperBound, whale)
  return(bestPos)
}
```

```{r}
# run WOA_chaotic
run_WOA_chaotic <- function(data) {
  # running WOA_chaotic
  sse_result <- WOA_chaotic(
    FUN = function(x) objective_SSE(x, data),
    optimType = "MIN",
    numVar = ncol(bounds),
    numPopulation = 40,
    maxIter = 500,
    rangeVar = bounds
  )
  
  # compute best SSE at that solution
  best_sse <- objective_SSE(sse_result, data)
  
  # return both
  list(
    h0 = sse_result[1],
    h1 = sse_result[2],
    t1 = sse_result[3],
    a  = sse_result[4],
    best_sse = best_sse
  )
}
```

## Quasirandom

```{r}
engineWOA_quasirandom <- function(FUN, optimType, maxIter, lowerBound, upperBound, whale) {
  whaleFitness <- calcFitness(FUN, optimType, whale)
  index <- order(whaleFitness)
  whaleFitness <- sort(whaleFitness)
  whale <- whale[index, ]
  bestPos <- whale[1, ]
  FbestPos <- whaleFitness[1]
  curve <- c()
  progressbar <- txtProgressBar(min = 0, max = maxIter, style = 3)

  # quasi-random sobol sequence
  dim_needed <- 2   # r1, r2
  sobol_seq <- sobol(n = maxIter, dim = dim_needed, scrambling = 0)

 
  # back to regular function
  for (t in 1:maxIter) {
    a <- 2 - t * ((2) / maxIter)
    a2 <- -1 + t * ((-1) / maxIter)

    for (i in 1:nrow(whale)) {

    r1 <- sobol_seq[t, 1]
    r2 <- sobol_seq[t, 2]
      

      A <- 2 * a * r1 - a
      C <- 2 * r2
      b <- 1
      p  <- runif(1)
      l  <- (a2 - 1) * runif(1) + 1


      for (j in 1:ncol(whale)) {
        if (p < 0.5) {
          if (abs(A) >= 1) {
            rand.index <- floor(nrow(whale) * r1) + 1            
            x.rand <- whale[rand.index, ]
            D.x.rand <- abs(C * x.rand[j] - whale[i, j])
            whale[i, j] <- x.rand[j] - A * D.x.rand
          } else {
            D.bestPos <- abs(C * bestPos[j] - whale[i, j])
            whale[i, j] <- bestPos[j] - A * D.bestPos
          }
        } else {
          distance <- abs(bestPos[j] - whale[i, j])
          whale[i, j] <- distance * exp(b * l) * cos(l * 2 * pi) + bestPos[j]
        }
      }

      whale[i, ] <- checkBound(whale[i, ], lowerBound, upperBound)
      fitness <- optimType * FUN(whale[i, ])
      if (fitness < FbestPos) {
        FbestPos <- fitness
        bestPos <- whale[i, ]
      }
    }
    curve[t] <- FbestPos
    setTxtProgressBar(progressbar, t)
  }
  close(progressbar)
  curve <- curve * optimType
  return(bestPos)
}

```


```{r}
WOA_quasirandom <- function(FUN, optimType = "MIN", numVar, numPopulation = 40, 
  maxIter = 500, rangeVar) 
{
  dimension <- ncol(rangeVar)
  lowerBound <- rangeVar[1, ]
  upperBound <- rangeVar[2, ]
  if (dimension == 1) {
    dimension <- numVar
  }
  if (optimType == "MAX") 
    optimType <- -1
  else optimType <- 1
  whale <- generateRandom(numPopulation, dimension, lowerBound, 
    upperBound)
  bestPos <- engineWOA_quasirandom(FUN, optimType, maxIter, lowerBound, 
    upperBound, whale)
  return(bestPos)
}
```

```{r}
run_WOA_quasirandom <- function(data) {
  # running WOA_quasirandom
  sse_result <- WOA_quasirandom(
    FUN = function(x) objective_SSE(x, data),
    optimType = "MIN",
    numVar = ncol(bounds),
    numPopulation = 40,
    maxIter = 500,
    rangeVar = bounds
  )
  
  # compute best SSE at that solution
  best_sse <- objective_SSE(sse_result, data)
  
  # return both
  list(
    h0 = sse_result[1],
    h1 = sse_result[2],
    t1 = sse_result[3],
    a  = sse_result[4],
    best_sse = best_sse
  )
}
```

## WOA

```{r}
run_WOA <- function(data) {
  # running WOA
  sse_result <- WOA(
    FUN = function(x) objective_SSE(x, data),
    optimType = "MIN",
    numVar = ncol(bounds),
    numPopulation = 40,
    maxIter = 500,
    rangeVar = bounds
  )
  
  # compute best SSE at that solution
  best_sse <- objective_SSE(sse_result, data)
  
  # return both
  list(
    h0 = sse_result[1],
    h1 = sse_result[2],
    t1 = sse_result[3],
    a  = sse_result[4],
    best_sse = best_sse
  )
}
```

## Levy

```{r}
levy_step <- function(n = 10000, beta = 1.5) {
  
  
  # same formula as above but lambda instead of beta
  sigma_u <- (gamma(1 + beta) * sin(pi * beta / 2) /
               (gamma((1 + beta) / 2) * beta * 2^((beta - 1) / 2)))^(1 / beta)
  
  # same u and v
  u <- rnorm(n, mean = 0, sd = sigma_u)
  v <- rnorm(n, mean = 0, sd = 1)
  
  # generate s
  s <- u / abs(v)^(1 / beta)
  
  return(s)
}
```

```{r}
engine_WOA_levy_at_end <- function(FUN, optimType, maxIter, lowerBound, upperBound, whale) {
  whaleFitness <- calcFitness(FUN, optimType, whale)
  index <- order(whaleFitness)
  whaleFitness <- sort(whaleFitness)
  whale <- whale[index, ]
  bestPos <- whale[1, ]
  FbestPos <- whaleFitness[1]
  curve <- c()
  progressbar <- txtProgressBar(min = 0, max = maxIter, style = 3)
  for (t in 1:maxIter) {
    a <- 2 - t * ((2)/maxIter)
    a2 <- -1 + t * ((-1)/maxIter)
    for (i in 1:nrow(whale)) {
      r1 <- runif(1)
      r2 <- runif(1)
      A <- 2 * a * r1 - a
      C <- 2 * r2
      b <- 1
      l <- (a2 - 1) * runif(1) + 1
      p <- runif(1)
      for (j in 1:ncol(whale)) {
        if (p < 0.5) {
          if (abs(A) >= 1) {
            rand.index <- floor(nrow(whale) * runif(1) + 
              1)
            x.rand <- whale[rand.index, ]
            D.x.rand <- abs(C * x.rand[j] - whale[i, 
              j])
            whale[i, j] <- x.rand[j] - A * D.x.rand
          }
          else if (abs(A) < 1) {
            D.bestPos <- abs(C * bestPos[j] - whale[i, 
              j])
            whale[i, j] <- bestPos[j] - A * D.bestPos
          }
        }
        else if (p >= 0.5) {
          distance <- abs(bestPos[j] - whale[i, j])
          whale[i, j] <- distance * exp(b * l) * cos(l * 
            2 * pi) + bestPos[j]
        }
      }
      whale[i, ] <- checkBound(whale[i, ], lowerBound, 
        upperBound)
      fitness <- optimType * FUN(whale[i, ])
      if (fitness < FbestPos) {
        FbestPos <- fitness
        bestPos <- whale[i, ]
      }
    }
    
# this is the levy part
    
scalar <- 0.01  # step-size scaling which we could tune...

for (i in 1:nrow(whale)) {
  # generate levy vector for each whale
  s <- levy_step(n = ncol(whale), beta = 1.5)
  
  # move whale by one levy step
  whale[i, ] <- whale[i, ] + scalar * s
  
  # check bounds
  whale[i, ] <- checkBound(whale[i, ], lowerBound, upperBound)
  
  # recalc fitness
  fitness <- optimType * FUN(whale[i, ])
  if (fitness < FbestPos) {
    FbestPos <- fitness
    bestPos <- whale[i, ]
  }
}
    # end of levy part
    curve[t] <- FbestPos
    setTxtProgressBar(progressbar, t)
  }
  close(progressbar)
  curve <- curve * optimType
  return(bestPos)
}
```


```{r}
WOA_levy_at_end <- function(FUN, optimType = "MIN", numVar, numPopulation = 40, 
  maxIter = 500, rangeVar) 
{
  dimension <- ncol(rangeVar)
  lowerBound <- rangeVar[1, ]
  upperBound <- rangeVar[2, ]
  if (dimension == 1) {
    dimension <- numVar
  }
  if (optimType == "MAX") 
    optimType <- -1
  else optimType <- 1
  whale <- generateRandom(numPopulation, dimension, lowerBound, 
    upperBound)
  bestPos <- engine_WOA_levy_at_end(FUN, optimType, maxIter, lowerBound, 
    upperBound, whale)
  return(bestPos)
}
```

```{r}
run_WOA_levy <- function(data) {
  # Run WOA_levy_at_end optimization
  sse_result <- WOA_levy_at_end(
    FUN = function(x) objective_SSE(x, data),
    optimType = "MIN",
    numVar = ncol(bounds),
    numPopulation = 40,
    maxIter = 500,
    rangeVar = bounds
  )
  
  # Compute best SSE for that solution
  best_sse <- objective_SSE(sse_result, data)
  
  # Return parameter estimates and SSE
  list(
    h0 = sse_result[1],
    h1 = sse_result[2],
    t1 = sse_result[3],
    a  = sse_result[4],
    best_sse = best_sse
  )
}
```


## LMA

```{r}
# first define SSE function
sse_func <- function(h0, h1, t1, a, data) {
  sse <- data |>
    mutate(preds = h0 + (h1 - h0) / (1 + exp(-a * (x - t1))),
      diff = (y - preds)^2) |>
    summarize(sse = sum(diff)) |>
    pull()
  return(sse) # i'd rather this be a vector, but this only way i can get it to work down below
}

# and sig_formula
sig_formula <- function(x, h0, h1, t1, a){
h0 + (h1 - h0) / (1 + exp(-a * (x - t1)))
}
```

```{r}
run_lma_fit <- function(data, dataset_id, seed) {
  # define bounds & starting values
  startlist <- list(h0 = 0.15, h1 = 0.75, t1 = 0.25, a = 90)
  lowerbounds <- c(h0 = -0.1, h1 = 0.3, t1 = -52, a = 0.01)
  upperbounds <- c(h0 = 0.35,  h1 = 0.97, t1 = 1, a = 180)
  
  # safely run LMA in case of convergence errors
  safe_nlsLM <- purrr::safely(minpack.lm::nlsLM, otherwise = NULL)
  
  result <- safe_nlsLM(
    intensity ~ sig_formula(time, h0, h1, t1, a),
    data = data,
    start = startlist,
    control = list(maxiter = 1000, factor = 100, ptol = 1e-8, ftol = 1e-8),
    lower = lowerbounds,
    upper = upperbounds,
    trace = FALSE
  )
  
  # If it failed, return NA row
  if (is.null(result$result)) {
    return(tibble(
      dataset_id = dataset_id,
      seed = seed,
      h0 = NA, h1 = NA, t1 = NA, a = NA,
      best_sse = NA,
      algorithm = "LMA"
    ))
  }
  
  # extract estimated parameters
  params <- coef(result$result)
  
  # compute SSE
  sse_val <- sse_func(
    h0 = params["h0"],            
    h1 = params["h1"],
    t1 = params["t1"],
    a  = params["a"],
    data = data |> rename(x = time, y = intensity)
  )
  
  # return as tibble
  tibble(
    h0 = params["h0"],
    h1 = params["h1"],
    t1 = params["t1"],
    a  = params["a"],
    best_sse = sse_val,
    dataset_id = dataset_id,
    seed = seed,
    algorithm = "LMA"
  )
}
```





# WOA double sigmoidal variants

## Chaotic

```{r}
run_WOA_chaotic_double <- function(data) {
  # running WOA_quasirandom
  sse_result <- WOA_chaotic(
    FUN = function(x) objective_SSE_double(x, data),
    optimType = "MIN",
    numVar = ncol(bounds),
    numPopulation = 40,
    maxIter = 500,
    rangeVar = bounds
  )
  
  # compute best SSE at that solution
  best_sse <- objective_SSE_double(sse_result, data)
  
  # return both
  list(
    h0 = sse_result[1],
    h1 = sse_result[2],
    h2 = sse_result[3],
    a1 = sse_result[4],
    a2 = sse_result[5],
    t1 = sse_result[6],
    t2 = sse_result[7],
    best_sse = best_sse
  )
}
```

## Quasirandom

```{r}
run_WOA_quasirandom_double <- function(data) {
  # running WOA_quasirandom
  sse_result <- WOA_quasirandom(
    FUN = function(x) objective_SSE_double(x, data),
    optimType = "MIN",
    numVar = ncol(bounds),
    numPopulation = 40,
    maxIter = 500,
    rangeVar = bounds
  )
  
  # compute best SSE at that solution
  best_sse <- objective_SSE_double(sse_result, data)
  
  # return both
  list(
    h0 = sse_result[1],
    h1 = sse_result[2],
    h2 = sse_result[3],
    a1 = sse_result[4],
    a2 = sse_result[5],
    t1 = sse_result[6],
    t2 = sse_result[7],
    best_sse = best_sse
  )
}
```

## WOA

```{r}
run_WOA_double <- function(data) {
  # running WOA_quasirandom
  sse_result <- WOA(
    FUN = function(x) objective_SSE_double(x, data),
    optimType = "MIN",
    numVar = ncol(bounds),
    numPopulation = 40,
    maxIter = 500,
    rangeVar = bounds
  )
  
  # compute best SSE at that solution
  best_sse <- objective_SSE_double(sse_result, data)
  
  # return both
  list(
    h0 = sse_result[1],
    h1 = sse_result[2],
    h2 = sse_result[3],
    a1 = sse_result[4],
    a2 = sse_result[5],
    t1 = sse_result[6],
    t2 = sse_result[7],
    best_sse = best_sse
  )
}
```

## Levy

```{r}
run_WOA_levy_double <- function(data) {
  # Run WOA_levy_at_end optimization
  sse_result <- WOA_levy_at_end(
    FUN = function(x) objective_SSE_double(x, data),
    optimType = "MIN",
    numVar = ncol(bounds),
    numPopulation = 40,
    maxIter = 500,
    rangeVar = bounds
  )
  
  # Compute best SSE for that solution
  best_sse <- objective_SSE_double(sse_result, data)
  
  # Return parameter estimates and SSE
  list(
    h0 = sse_result[1],
    h1 = sse_result[2],
    h2 = sse_result[3],
    a1 = sse_result[4],
    a2 = sse_result[5],
    t1 = sse_result[6],
    t2 = sse_result[7],
    best_sse = best_sse
  )
}
```


## LMA

```{r}
# double sigmoid lma
run_lma_fit_double <- function(data, dataset_id, seed, 
                                      max_success = 20, max_attempts = 500) {
  set.seed(seed)
  
  # bounds
  lowerbounds <- c(h0 = -0.10, h1 = 0.30, h2 = 0.00,
                   a1 = 0.01, a2 = 0.01,
                   t1 = -0.52, t2 = -0.48)
  upperbounds <- c(h0 = 0.30, h1 = 0.97, h2 = 1.00,
                   a1 = 180.00, a2 = 180.00,
                   t1 = 1.15, t2 = 1.79)
  
  # safe nlsLM
  safe_nlsLM <- purrr::safely(minpack.lm::nlsLM, otherwise = NULL)
  
  # store successful fits
  fits <- list()
  sse_values <- c()
  attempts <- 0
  
  while(length(fits) < max_success && attempts < max_attempts) {
    attempts <- attempts + 1
    
    # random start within bounds
    startlist <- lapply(names(lowerbounds), function(nm) {
      runif(1, min = lowerbounds[nm], max = upperbounds[nm])
    })
    names(startlist) <- names(lowerbounds)
    
    # make sure t2 > t1
    if(startlist$t2 <= startlist$t1) next
    
    # attempt nlsLM
    res <- safe_nlsLM(
      intensity ~ double_sig_fit_formula(time, h0, h1, h2, a1, a2, t1, t2),
      data = data,
      start = startlist,
      lower = lowerbounds,
      upper = upperbounds,
      control = list(maxiter = 1000, factor = 100, ptol = 1e-8, ftol = 1e-8),
      trace = FALSE
    )
    
    if(!is.null(res$result)) {
      # compute SSE
      params <- coef(res$result)
      sse_val <- objective_SSE_double(unname(params), data)
      
      fits[[length(fits)+1]] <- params
      sse_values <- c(sse_values, sse_val)
    }
  }
  
  if(length(fits) == 0) {
    # no successful fits
    return(tibble(
      dataset_id = dataset_id,
      seed = seed,
      h0 = NA, h1 = NA, h2 = NA,
      a1 = NA, a2 = NA,
      t1 = NA, t2 = NA,
      best_sse = NA,
      algorithm = "LMA_random"
    ))
  }
  
  # choose best fit (lowest SSE)
  best_idx <- which.min(sse_values)
  best_params <- fits[[best_idx]]
  
  tibble(
    dataset_id = dataset_id,
    seed = seed,
    h0 = best_params["h0"],
    h1 = best_params["h1"],
    h2 = best_params["h2"],
    a1 = best_params["a1"],
    a2 = best_params["a2"],
    t1 = best_params["t1"],
    t2 = best_params["t2"],
    best_sse = sse_values[best_idx],
    algorithm = "LMA_random"
  )
}

```

LMA without random starts

```{r}
# this is just to keep the old one - it shouldn't be used anywhere
# double sigmoid lma
run_lma_fit_double_no_random <- function(data, dataset_id, seed) {
  
   # define bounds & starting values
  startlist <- list(h0 = 0.15, h1 = 0.75, h2 = 0.5, a1 = 90, a2 = 90, t1 = 0.25, t2 = 0.3)
  lowerbounds <- c(h0 = -0.10, h1 = 0.30, h2 = 0.00, a1 = 0.01, a2 = 0.01, t1 = -0.52, t2 = -0.48)
  upperbounds <- c(h0 = 0.30, h1 = 0.97, h2 = 1.00, a1 = 180.00, a2 = 180.00, t1 = 1.15, t2 = 1.79)
  
  
  # safely run LMA in case of convergence errors
  safe_nlsLM <- purrr::safely(minpack.lm::nlsLM, otherwise = NULL)
  
  result <- safe_nlsLM(
    intensity ~ double_sig_fit_formula(time, h0, h1, h2, a1, a2, t1, t2),
    data = data,
    start = startlist,
    control = list(maxiter = 1000, factor = 100, ptol = 1e-8, ftol = 1e-8),
    lower = lowerbounds,
    upper = upperbounds,
    trace = FALSE
  )
  
  # If it failed, return NA row
  if (is.null(result$result)) {
    return(tibble(
      dataset_id = dataset_id,
      seed = seed,
      h0 = NA, h1 = NA, h2 = NA,
      a1 = NA, a2 = NA,
      t1 = NA, t2 = NA,
      best_sse = NA,
      algorithm = "LMA"
    ))
  }
  
  # extract estimated parameters
  params <- coef(result$result)
  
 # compute SSE
params_vec <- c(
  params["h0"], params["h1"], params["h2"],
  params["a1"], params["a2"], params["t1"], params["t2"]
)

sse_val <- objective_SSE_double(params_vec, data)

  
  # return as tibble
  tibble(
    h0 = params["h0"],
    h1 = params["h1"],
    h2 = params["h2"],
    a1 = params["a1"],
    a2 = params["a2"],
    t1 = params["t1"],
    t2 = params["t2"],
    best_sse = sse_val,
    dataset_id = dataset_id,
    seed = seed,
    algorithm = "LMA"
  )
}


```



# Data Generation

Sigmoidal

```{r}
data_gen_func <- function(h0 = 30, h1 = 750, t1 = 160, a = 0.05, reps = 5, disp = 250){
  

sig_formula <- function(x, h0, h1, t1, a){
h0 + (h1 - h0) / (1 + exp(-a * (x - t1)))
}

x_vals <- rep(seq(0, 300, by = 20), times = reps)

true_params <- list(h0 = h0, h1 = h1, t1 = t1, a = a)

mean_vals <- rep(sig_formula(x_vals,
                        h0 = true_params$h0,
                        h1 = true_params$h1,
                        t1 = true_params$t1,
                        a = true_params$a))

data <- rnorm(n = length(mean_vals), mean = mean_vals, sd = rep(disp, length(mean_vals)))
sig_data <- data.frame(time = x_vals, intensity = data)

# also want a normalized version
norm_sig_data <- sig_data |>
  mutate(time = ((time - min(time))/(max(time)- min(time))),
         intensity = ((intensity - min(intensity))/(max(intensity)- min(intensity))))
}
```

Double sigmoidal

```{r}
# generate data
generate_piecewise_data <- function(x_vals = seq(0, 300, by = 20), reps = 5, dispersion = 10,
    a1 = 0.04, a2 = 0.03, t1 = 110, t2 = 240, h1 = 750, h2 = 400, h0 = 20) {
 

  # set.seed(101)
  
  # find x-value corresponding to max of f0 (which is t*)
  x_dense <- seq(min(x_vals), max(x_vals), length.out = 1000)
  f0_vals <- f0(x_dense, a1, t1, a2, t2)
  argument <- x_dense[which.max(f0_vals)]

  # const = f0
  const <- f0(argument, a1, t1, a2, t2) # argument because max(f0) is at t*

  # add reps to x and y
  x_repeated <- rep(x_vals, each = reps)
  y_true <- f2_h0(x_repeated, h0, h1, h2, a1, t1, a2, t2, const, argument)
  y_noisy <- rnorm(length(y_true), mean = y_true, sd = dispersion)

  # make df
  df <- data.frame(
    time = x_repeated,
    intensity = y_noisy)
    
  norm_df <- df |>
  mutate(time = ((time - min(time))/(max(time)- min(time))),
         intensity = ((intensity - min(intensity))/(max(intensity)- min(intensity))))
  
  return(norm_df)
  
}
```



# Running stuff

```{r}
# set up multicores
plan(multisession, workers = 5)  # i have 8 total


tic()
seeds <- 1:100

# disp = 250

# generate the data with given seeds
generated_data <- map(seeds, function(s) {
  set.seed(s)
  data_gen_func()
})

# define bounds
bounds <- matrix(c(-0.1, 0.3,
                   -0.52, 0.01,
                   0.35, 0.97,
                   1, 180),
                 nrow = 2, ncol = 4, byrow = TRUE)

# helper function to run everything smoothly in parallel
run_woa_parallel <- function(data, i, algo_func, algo_name) {
  out <- withSeed({
    algo_func(data)
  }, seed = seeds[i])  # ensures all random numbers inside algo_func are reproducible

  out$dataset_id <- i
  out$seed <- seeds[i]
  out$algorithm <- algo_name
  out
}

# og WOA
og_df_250 <- future_map2_dfr(
  generated_data, seq_along(generated_data),
  ~ run_woa_parallel(.x, .y, run_WOA, "WOA"),
  seed = TRUE  # ensures parallel-safe independent RNG streams
)

# chaotic WOA
chaotic_df_250 <- future_map2_dfr(
  generated_data, seq_along(generated_data),
  ~ run_woa_parallel(.x, .y, run_WOA_chaotic, "WOA_chaotic"),
  seed = TRUE
)

# quasirandom WOA
quasi_df_250 <- future_map2_dfr(
  generated_data, seq_along(generated_data),
  ~ run_woa_parallel(.x, .y, run_WOA_quasirandom, "WOA_quasirandom"),
  seed = TRUE
)

# levy WOA
levy_df_250 <- future_map2_dfr(
  generated_data, seq_along(generated_data),
  ~ run_woa_parallel(.x, .y, run_WOA_levy, "WOA_levy"),
  seed = TRUE
)

lma_results_df_250 <- map2_dfr(generated_data, seeds, ~run_lma_fit(.x, dataset_id = which(seeds == .y), seed = .y))

# combine
all_results_250 <- bind_rows(og_df_250, chaotic_df_250, quasi_df_250, levy_df_250, lma_results_df_250)


# disp = 400

# generate data
generated_data_400 <- map(seeds, function(s) {
  set.seed(s)
  data_gen_func(disp = 400) # change dispersion here
})

# define bounds
bounds <- matrix(c(-0.1, 0.3,
                   -0.52, 0.01,
                   0.35, 0.97, # this should maybe be 0.3 lower bound for h1
                   1, 180),
                 nrow = 2, ncol = 4, byrow = TRUE)


# og WOA
og_df_400 <- future_map2_dfr(
  generated_data_400, seq_along(generated_data_400),
  ~ run_woa_parallel(.x, .y, run_WOA, "WOA"),
  seed = TRUE  # ensures parallel-safe independent RNG streams
)

# chaotic WOA
chaotic_df_400 <- future_map2_dfr(
  generated_data_400, seq_along(generated_data_400),
  ~ run_woa_parallel(.x, .y, run_WOA_chaotic, "WOA_chaotic"),
  seed = TRUE
)

# quasirandom WOA
quasi_df_400 <- future_map2_dfr(
  generated_data_400, seq_along(generated_data_400),
  ~ run_woa_parallel(.x, .y, run_WOA_quasirandom, "WOA_quasirandom"),
  seed = TRUE
)

# levy WOA
levy_df_400 <- future_map2_dfr(
  generated_data_400, seq_along(generated_data_400),
  ~ run_woa_parallel(.x, .y, run_WOA_levy, "WOA_levy"),
  seed = TRUE
)

lma_results_df_400 <- map2_dfr(generated_data_400, seeds, ~run_lma_fit(.x, dataset_id = which(seeds == .y), seed = .y))


# combine
all_results_400 <- bind_rows(og_df_400, chaotic_df_400, quasi_df_400, levy_df_400, lma_results_df_400)


# double sigmoidal

# generate the data with given seeds
double_sig_data <- map(seeds, function(s) {
  set.seed(s)
  generate_piecewise_data()
})

# define bounds
bounds <- matrix(c(-0.10,  0.30, 0.00,   0.01,   0.01,  -0.52, -0.48,
                   0.30,  0.97, 1.00, 180.00, 180.00,   1.15,  1.79), nrow = 2, byrow = TRUE)



# og WOA
double_og_df <- future_map2_dfr(
  double_sig_data, seq_along(double_sig_data),
  ~ run_woa_parallel(.x, .y, run_WOA_double, "WOA"),
  seed = TRUE  # ensures parallel-safe independent RNG streams
)

# chaotic WOA
double_chaotic_df <- future_map2_dfr(
  double_sig_data, seq_along(double_sig_data),
  ~ run_woa_parallel(.x, .y, run_WOA_chaotic_double, "WOA_chaotic"),
  seed = TRUE
)

# quasirandom WOA
double_quasi_df <- future_map2_dfr(
  double_sig_data, seq_along(double_sig_data),
  ~ run_woa_parallel(.x, .y, run_WOA_quasirandom_double, "WOA_quasirandom"),
  seed = TRUE
)

# levy WOA
double_levy_df <- future_map2_dfr(
  double_sig_data, seq_along(double_sig_data),
  ~ run_woa_parallel(.x, .y, run_WOA_levy_double, "WOA_levy"),
  seed = TRUE
)


lma_double_results_df <- map2_dfr(
  double_sig_data,  
  seeds,
  ~run_lma_fit_double(.x, dataset_id = which(seeds == .y), seed = .y)
)

# combine
all_results_double <- bind_rows(double_og_df, double_chaotic_df, double_quasi_df, double_levy_df, lma_double_results_df)
toc()
```

It took ~6.5 hours to run on 100 datasets.

# Plots

```{r}
all_results_250 |>
  ggplot(aes(x = h0, y = t1)) +
  geom_point(aes(color = algorithm)) +
  geom_hline(yintercept = 0.53, color = "red") +
  geom_vline(xintercept = 0.15, color = "red") +
  facet_wrap(~algorithm)
```

```{r}
all_results_250 |>
  ggplot(aes(x = a, y = h1)) +
  geom_point(aes(color = algorithm)) +
  geom_hline(yintercept = 0.83, color = "red") +
  geom_vline(xintercept = 15, color = "red") +
  facet_wrap(~algorithm)
```

```{r}
# disp 250
lowest_sse_df_250 <- all_results_250 |>
  group_by(dataset_id) |>
  mutate(lowest_sse = min(best_sse)) |>
  filter(best_sse == lowest_sse) |>
  select(dataset_id, algorithm, lowest_sse)

lowest_sse_df_250 |>
  group_by(algorithm) |>
  summarise(count = n()) |>
  arrange(desc(count))

# disp 400
lowest_sse_df_400 <- all_results_400 |>
  group_by(dataset_id) |>
  mutate(lowest_sse = min(best_sse)) |>
  filter(best_sse == lowest_sse) |>
  select(dataset_id, algorithm, lowest_sse)

lowest_sse_df_400 |>
  group_by(algorithm) |>
  summarise(count = n()) |>
  arrange(desc(count))

```


Double is interesting. The SSE values to not match the plot.

```{r}
# double
lowest_sse_df_double <- all_results_double |>
  group_by(dataset_id) |>
  mutate(lowest_sse = min(best_sse)) |>
  filter(best_sse == lowest_sse) |>
  select(dataset_id, algorithm, lowest_sse)

lowest_sse_df_double |>
  group_by(algorithm) |>
  summarise(count = n()) |>
  arrange(desc(count))

n_distinct(lowest_sse_df_double$dataset_id)

```

This was code to find true values in normal space...  
  
```{r}
x_vals = seq(0, 300, by = 20)
reps = 5
dispersion = 10
a1 = 0.04
a2 = 0.03
t1 = 110
t2 = 240
h1 = 750
h2 = 400
h0 = 20
 

  # set.seed(101)
  
  # find x-value corresponding to max of f0 (which is t*)
  x_dense <- seq(min(x_vals), max(x_vals), length.out = 1000)
  f0_vals <- f0(x_dense, a1, t1, a2, t2)
  argument <- x_dense[which.max(f0_vals)]

  # const = f0
  const <- f0(argument, a1, t1, a2, t2) # argument because max(f0) is at t*

  # add reps to x and y
  x_repeated <- rep(x_vals, each = reps)
  y_true <- f2_h0(x_repeated, h0, h1, h2, a1, t1, a2, t2, const, argument)
  y_noisy <- rnorm(length(y_true), mean = y_true, sd = dispersion)

  # make df
  df <- data.frame(
    time = x_repeated,
    intensity = y_noisy)
```
  

```{r}
# checking normalization
# h1
h0_norm_func <- function(data, h0) {
  h0_norm <- (h0 - min(data$intensity))/(max(data$intensity) - min(data$intensity))
  return(h0_norm)
}


# h1
h1_norm_func <- function(data, h1) {
  h1_norm <- (h1 - min(data$intensity))/(max(data$intensity) - min(data$intensity))
  return(h1_norm)
}

# t1
t1_norm_func <- function(data, t1) {
  t1_norm <- (t1 - min(data$time))/(max(data$time) - min(data$time))
  return(t1_norm)
}

# a
a_norm_func <- function(data, a) {
  a_norm <- a * (max(data$time) - min(data$time))
  return(a_norm)
}

# h0, h1, h2
h0_norm_func(df, h0)
h1_norm_func(df, h1)
h1_norm_func(df, h2)

# t1, t2
t1_norm_func(df, M1)
t1_norm_func(df, M1 + L)

# a1, a2
a_norm_func(df, B1)
a_norm_func(df, B2)

```

```{r}
all_results_double |>
  ggplot(aes(x = h0, y = h1)) +
  geom_point(aes(color = algorithm)) +
  geom_hline(yintercept = .96, color = "red") +
  geom_vline(xintercept = -0.014, color = "red") +
  facet_wrap(~algorithm)
```


```{r}
all_results_double |>
  ggplot(aes(x = t1, y = t2)) +
  geom_point(aes(color = algorithm)) +
  geom_hline(yintercept = 0.8, color = "red") +
  geom_vline(xintercept = 0.3666667, color = "red") +
  facet_wrap(~algorithm)
```

When you look at the data, you can see there are a lot of fits where t1 and t2 are at their respective lower bounds...

```{r}
all_results_double |>
  ggplot(aes(x = a1, y = a2)) +
  geom_point(aes(color = algorithm)) +
  geom_hline(yintercept = 9, color = "red") +
  geom_vline(xintercept = 12, color = "red") +
  facet_wrap(~algorithm)
```


```{r}
all_results_double |>
  ggplot(aes(x = h1, y = h2)) +
  geom_point(aes(color = algorithm)) +
  geom_hline(yintercept = 0.49, color = "red") +
  geom_vline(xintercept = 0.96, color = "red") +
  facet_wrap(~algorithm)
```

```{r}
# want to get a sense of difference between sse of each algorithm - is 
# WOA_levy closer to lma or the others
all_results_double |>
  ggplot(aes(x = reorder(algorithm, best_sse), y = best_sse)) +
  geom_boxplot() +
  labs(x = "Algorithm", y = "SSE", title = "Distribution of SSE Across Algorithms") +
  theme_minimal()

# with just means
mean_sse_by_alg <- all_results_double |>
  group_by(algorithm) |>
  summarise(mean_sse = mean(best_sse, na.rm = TRUE),
    sd_sse = sd(best_sse, na.rm = TRUE),
    n = n()) |>
  arrange(mean_sse)

mean_sse_by_alg |>
  ggplot(aes(x = reorder(algorithm, mean_sse), y = mean_sse)) +
  geom_segment(aes(x = reorder(algorithm, mean_sse),
    xend = reorder(algorithm, mean_sse),
    y = 0,
    yend = mean_sse)) +
  geom_point(size = 4, color = "violet") +
  labs(x = "Algorithm", y = "Mean SSE") +
  theme_minimal()


  
```



# Testing fits

```{r}
# set seed for a dataset that seemed reasonable
set.seed(1)
double_test_data_1 <- generate_piecewise_data()

double_test_data_1 |>
  ggplot(aes(x = time, y = intensity)) +
  geom_point() +
  geom_line(aes(x = time, y = double_sig_fit_formula(time, h0 = 0.105268913, h1 = 0.8919080, h2 = 0.6262562, a1 = 75.69219, a2 = 73.13922, t1 = 0.3349718, t2 = 0.8583771), color = "WOA_levy")) +
  geom_line(aes(x = time, y = double_sig_fit_formula(time, h0 = 0.032350751, h1 = 0.9700000, h2 = 0.1888878, a1 = 13.85791, a2 = 5.04554, t1 = 0.3602568, t2 = 0.9289089), color = "WOA")) +
  geom_line(aes(x = time, y = double_sig_fit_formula(time, h0 = 0.115041336, h1 = 0.8200104, h2 = 0.3689996, a1 = 83.75139, a2 = 78.80577, t1 = 0.3327068, t2 = 1.5435706), color = "WOA_chaotic")) +
  geom_line(aes(x = time, y = double_sig_fit_formula(time, h0 = 0.083572760, h1 = 0.8457848, h2 = 0.3257005, a1 = 83.06194, a2 = 75.86892, t1 = 0.3327265, t2 = 1.0005449), color = "WOA_quasirandom")) +
  geom_line(aes(x = time, y = double_sig_fit_formula(time, h0 = 0.007842110, h1 = 0.9700000, h2 = 0.5180706, a1 = 13.20930, a2 = 10.27320, t1 = 0.3510049, t2 = 0.8309659), color = "LMA_random")) +
  labs(title = "Fits on dataset 1") + 
  geom_hline(yintercept = 0.97, color = "red" , linetype = "dashed")
  


```













This is just testing a single dataset making sure stuff works

```{r}
# generate a dataset
set.seed(47)
double_test_data <- generate_piecewise_data()


run_WOA_double(double_test_data)
run_WOA_chaotic_double(double_test_data)
run_WOA_quasirandom_double(double_test_data)
run_WOA_levy_double(double_test_data)
# run_lma_fit_double_random(double_test_data, dataset_id = 47, seed = 47)
run_lma_fit_double(double_test_data, dataset_id = 47, seed = 47)

```



